---
title: "A1: Getting started — your first discrimination report card"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  gradepath turns firm-level empirical-Bayes estimates into a discrimination
  *report card*: it ranks units by a posterior outranking probability, solves a
  grade integer program along an information-reliability frontier, and assembles
  a grade-plus-interval table you can print, plot, and export. This vignette is a
  gentle first contact. In under 60 lines of code you install the package, see the
  one `krw_report_card()` call you would write, load the bundled RACE report card
  (97 firms, grades 2 / 81 / 14), and read it three ways: as a console box, as a
  tidy data frame, and as one figure. You leave knowing what input a report card
  needs, how to read its grades and reliability honestly, and where the full
  workflow lives next.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A1: Getting started — your first discrimination report card}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. Why you are here

You have a vector of unit-level estimates — say, 97 discrimination point
estimates across large U.S. firms — each with its own standard error. The
noisiest estimates look the most extreme purely by chance, so a raw ordering
mistakes imprecision for evidence. How do you sort units by *what the data
actually support*, and report each one with an honest interval, without
overstating a difference the noise cannot bear?

> "Estimates of firm-level discrimination are inherently noisy ... A report
> card summarizes this evidence by sorting firms into a small number of grades."
> — Kline, Rose & Walters [@kline2024discrimination]

gradepath is a native-R replication companion for that paper. This vignette runs
the full report-card workflow in one conceptual call, then unpacks what just
happened.

## 2. What you will leave with

After this vignette, you will be able to:

- Install **gradepath** and load a bundled report card with one call.
- Read the package's console box and locate the grade composition, the
  reliability `(1 - DR)`, and `tau-bar`.
- Pull the report card into your own workflow as a tidy data frame and export it
  to CSV.
- Read the one signature figure — each firm's posterior gap with its interval,
  ranked and coloured by grade.
- Understand, in plain steps, the three-part KRW pipeline (estimate → rank →
  grade) without touching a derivation.

Reading + reflection ≈ 25–35 min. It runs in seconds, because it loads a
pre-computed fit rather than solving anything.

## 3. Install and load

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("joonho112/gradepath")
```

The setup chunk above already attached **gradepath** and **ggplot2** and set the
package theme, so you do not need to load anything else. For your own scripts the
one line to remember is:

```{r library, eval = FALSE}
library(gradepath)
```

## 4. The input — what a report card needs

A report card needs very little: a vector of per-firm gaps and a vector of their
standard errors. Each gap is one firm's estimated discrimination (here, the
White-minus-Black callback gap); each standard error says how precisely that gap
was measured.

The package ships a public example, `krw_firms`, that carries exactly this shape
(97 firms with race and gender gaps and SEs). It is handy for trying the API, but
it is an example on a *different numeric scale* and does **not** reproduce the
paper's grades — so we do not drive the headline off it. The figures in this
vignette come from the bundled parity input: KRW's real per-firm beta-GMM series,
shipped under `inst/extdata/krw-gmm-input/`, which is what reproduces the
published RACE result.

## 5. The one call

The whole workflow is one function, `krw_report_card()`. You assemble the input
from the shipped beta-GMM series and hand it over with a control:

```{r one-call, eval = FALSE}
# Read the real per-firm beta-GMM series shipped with the package (97 firms,
# headerless: column 2 = theta_hat gap, column 3 = s standard error).
gmm  <- read.csv(system.file("extdata/krw-gmm-input/theta_estimates_matlab_race.csv",
                             package = "gradepath"), header = FALSE)
race <- list(theta_hat = gmm[[2]], s = gmm[[3]])

fit <- krw_report_card(
  race, demographic = "race",
  control = gp_control(backend = "gurobi", precision_rule = "krw_gmm",
                       lambda_grid = c(0.25, 0.5, 1)))   # the grid must include 1
```

That single call runs all seven stages in order: beta-GMM precision
standardization → deconvolution of the prior → posterior shrinkage → the
pairwise outranking matrix → the grade integer program → the
information-reliability frontier → the assembled report card.

The catch is the fifth stage. The grade step **solves an integer program over 97
firms**, which takes roughly three minutes with Gurobi. So we do not run it in
this vignette — gradepath ships the proven result, and we load it here:

```{r load-fit}
fit <- gp_parity_fit("race")
```

This bundled object is exactly what the call above produces: the RACE report card
with grades 2 / 81 / 14 at the KRW baseline `lambda = 0.25`. Everything below
reads off it; nothing below re-solves.

## 6. Read it: the console

The fastest read is to print the fit. The console box is result-first: how many
units, the grade composition, the selected `lambda`, the reliability, and the
backend.

```{r print-fit}
fit
```

What to look at:

- **`grades`** — `3 (2/81/14)`: 97 firms sorted into three grades, with 2 in the
  most-extreme grade, 81 in the middle, and 14 in the least-extreme grade.
- **`reliability: (1 - DR)`** — `0.96`: the share of comparable firm pairs on
  which the grades agree with the underlying posterior ranking. Higher is
  cleaner.
- **`tau-bar`** — `0.21`: the posterior expected Kendall rank agreement the
  grading carries across grade boundaries (between-grade blocks), in [-1, 1];
  higher means the grades preserve more rank information.

For the plain-English read, call `summary()`. On top of the same numbers it adds
an **Interpretation** block (what the grades, reliability, and `tau-bar` mean in
words), a **Glossary** (each term defined), and a provenance block (backend and
selection rule):

```{r summary-fit}
summary(fit)
```

The Interpretation block states the headline directly: the grades agree with the
pairwise (posterior) ranking on 96% of comparable pairs. Grades are a sorting by
posterior evidence, never a contest, and grade 1 is not a verdict — it is simply
the most-extreme grade under the data.

## 7. Read it: the table

For joins, regressions, or plots in your own style, pull the report card into a
data frame. `as.data.frame(fit)` returns a tidy 97 × 10 table, ranked with the
most-extreme unit (grade 1) first:

```{r as-dataframe}
head(as.data.frame(fit), 8)
```

The columns are `id`, `label`, `grade`, `sort_rank`, `selected_lambda`,
`posterior_mean`, `lower`, `upper`, `estimate`, and `se`. Compare `estimate` (the
raw gap) with `posterior_mean` (the shrunken gap) to watch empirical Bayes work:
the noisier a firm's estimate, the more its posterior mean is pulled toward the
center. The interval `[lower, upper]` is the credible interval carried on the
card; it is reported as-is and never recomputed downstream.

Two one-line reads confirm the headline numbers:

```{r reads}
round(get_prior(fit)$metadata$beta, 4)   # the precision exponent
table(get_grades(fit))                    # the grade composition
```

The precision exponent is $\beta$ in $\theta = s^{\beta}\, v$ — gradepath lands on
`0.5095`, matching the `0.510` KRW report. The grade table is the `2 / 81 / 14`
composition again.

Exporting is a one-liner:

```{r export, eval = FALSE}
write.csv(as.data.frame(fit), "report_card_race.csv", row.names = FALSE)
```

## 8. Read it: one figure

The single most informative figure is the report card itself (KRW Figure 7): each
firm's posterior gap shown as a point with its interval, one firm per row, ranked
with the most-extreme firm on top and coloured by grade.

```{r figure-report-card, fig.width = 7, fig.height = 8, fig.cap = "RACE report card (KRW Figure 7): each of the 97 firms shown as its posterior White-minus-Black callback gap (point) with its credible interval (horizontal line), one firm per row, ranked with the most-extreme firm on top and coloured by its grade (2 / 81 / 14).", fig.alt = "A horizontal caterpillar plot of 97 firms, one per row, ranked from most extreme at the top to least extreme at the bottom. Each row shows a point for the firm's posterior gap and a horizontal line for its credible interval. Points are coloured by grade into three groups: a small most-extreme grade at the top, a large middle grade, and a least-extreme grade at the bottom, in the 2 / 81 / 14 split."}
gp_plot_report_card(get_report_card(fit))
```

Reading the figure: the vertical position is the firm's rank; the horizontal
position is the posterior gap; the line length is the firm's uncertainty. The
colour blocks are the grades. Because the figure shows each interval, you can see
at a glance which firms are separated by the data and which overlap — the grades
respect that separation rather than the raw point order.

## 9. What you just did, formally

The KRW pipeline is three plain steps, no equations required.

1. **Estimate.** Take each firm's raw gap and standard error and put them on a
   common precision scale via the beta-GMM standardization (the precision
   exponent $\beta$ above).
2. **Rank.** Deconvolve the prior distribution of true gaps, shrink each firm's
   estimate toward it to get a posterior, and compare every pair of firms by the
   posterior probability that one outranks the other.
3. **Grade.** Solve an integer program that sorts firms into a small number of
   grades along the information-reliability frontier, selecting the grade count
   at the KRW baseline $\lambda = 0.25$.

Grades are a sorting by posterior evidence, not a ranking contest: a grade label
is an integer in `1 … n` that carries no superiority statement. The one-level
race and gender core is accepted within its declared scope, and the numbers here
match the paper ($\beta = 0.5095$ vs the reported `0.510`; $\lambda = 0.25$ is the
KRW baseline).

## 10. Where to next

- **Full workflow** — `r vlink("a2-the-grading-workflow", "the grading workflow")`
  walks the seven stages individually and shows race and gender side by side.
- **Reading and exporting** — `vignette("a3-reading-and-exporting-results")` goes deeper
  on the accessors, the table, and CSV/round-trip export.
- **Figures** — `vignette("a4-figures")` builds all four KRW
  signature figures.
- **Solvers and calibration** — `vignette("a5-solvers-and-calibration")` covers
  the open-source backends and the calibration harness.
- **Foundations** — `vignette("m1-foundations-and-notation")` formalizes the
  estimate → rank → grade pipeline with the math this vignette deliberately skipped.

### Provenance

```{r provenance}
sessionInfo()
```

## References
