---
title: "A2: The grading workflow — from estimates to a report card"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  In A1 (getting started), one `krw_report_card()` call ran the
  whole KRW pipeline and printed a finished report card. This vignette opens the
  box. We load the bundled RACE fit once and walk the seven stages one at a time,
  in order — precision standardization → prior deconvolution → posterior
  shrinkage → pairwise outranking → the grade integer program → the
  information-reliability frontier → the assembled report card — reading each
  stage's output off the loaded fit as a call plus its read. Only the grade step
  solves an integer program, so we show those calls but read the proven result
  instead of re-solving. At the end we recover the one-call shortcut and run the
  identical workflow for gender.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A2: The grading workflow — from estimates to a report card}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. From one call to seven stages

`r vlink("a1-getting-started", "A1")` ran the entire pipeline in a single line:

```{r recap-one-call, eval = FALSE}
fit <- krw_report_card(
  race, demographic = "race",
  control = gp_control(backend = "gurobi", precision_rule = "krw_gmm",
                       lambda_grid = c(0.25, 0.5, 1)))
```

That call hides seven KRW stages [@kline2024discrimination]. This vignette opens
the box and walks them in order, one stage per section. Each section has the same
shape: the call you would write, and the read of what it produces.

We load the proven RACE fit **once**, here, and read every stage off it. Loading
is instant because the fit is shipped pre-solved; nothing below re-solves.

```{r load-fit}
fit <- gp_parity_fit("race")
fit
```

The console box is the destination: 97 firms sorted into three grades
(`2 / 81 / 14`) at the KRW baseline `lambda = 0.25`, with reliability `0.96` and
`tau-bar = 0.21`. The rest of this vignette is the road there.

> A note on speed and honesty. Reading a finished stage off the loaded fit is
> fast, so those chunks are **evaluated**. The only functions that *solve* the
> grade integer program — `gp_grade()`, `gp_grade_path()`, and
> `krw_report_card()` — take minutes on 97 firms, so they appear as
> `eval = FALSE` code you could run, and we read the already-solved result
> instead.

## 2. The input

A report card needs only two per-firm vectors: the estimated gap and its standard
error. For RACE that is the White-minus-Black callback gap and its SE, one pair
per firm. The headline RACE result is driven by KRW's real per-firm beta-GMM
series, shipped under `inst/extdata/krw-gmm-input/`:

```{r input-recap, eval = FALSE}
gmm  <- read.csv(system.file("extdata/krw-gmm-input/theta_estimates_matlab_race.csv",
                             package = "gradepath"), header = FALSE)
race <- list(theta_hat = gmm[[2]], s = gmm[[3]])   # 97 firms: gap + SE
```

The public `krw_firms` dataset carries the same *shape* and is handy for trying
the API, but it is an example on a different numeric scale and does **not**
reproduce the published `2 / 81 / 14`. The loaded `fit` already carries this input
on its `estimates` slot, so every read below works directly off `fit`.

## 3. Stage 1 — precision standardization (beta-GMM)

The first stage puts every firm's gap on a common precision scale. KRW model the
observed estimate as $\theta = s^{\beta} v$, so the standard error $s$ enters with
an estimated exponent $\beta$ rather than being assumed away. That exponent is the
beta-GMM estimate; it lives in the prior's metadata:

```{r stage1-beta}
pr <- get_prior(fit)
round(pr$metadata$beta, 4)        # the precision exponent beta
```

gradepath lands on $\beta = 0.5095$, matching the `0.510` KRW report. A value near
$0.5$ says precision scales close to the square root of $s$: noisier firms are
standardized more strongly before any comparison is made. The rest of
`pr$metadata` (the characteristic, carrier count, support points, and the
deconvolution's `alpha`) records the stage's bookkeeping.

## 4. Stage 2 — deconvolving the prior

The same `pr` object carries the second stage's output. From the standardized
estimates the pipeline *deconvolves* the prior distribution $G$ of true gaps — it
backs out the distribution of latent $v$ that, blurred by sampling noise, would
reproduce the spread of observed estimates. The prior's shape is flexibly
estimated from the data rather than imposed. The recovered prior is a grid
(`support`) carrying a mass (`density`):

```{r stage2-prior}
prior_df <- data.frame(support = pr$support, density = pr$density)
head(prior_df)
pr$scale                                  # scale label: "r" (the r-scale), not a number
w      <- pr$density / sum(pr$density)    # normalise the grid mass
spread <- sqrt(sum(w * (pr$support - pr$mean)^2))  # density-weighted SD of the prior
c(centre = pr$mean, spread = spread)
```

`pr$mean` is the centre of that recovered prior, while `pr$scale` is the scale
label (here `"r"`, the r-scale) rather than a spread; the spread is the
density-weighted standard deviation of the `support` grid (the `spread` computed
in the chunk above). This
distribution is what the next stage shrinks toward: it encodes how much true
variation across firms the data actually support, separated from noise.

## 5. Stage 3 — posterior shrinkage

With the prior in hand, the third stage produces each firm's **posterior**: it
combines the firm's own noisy estimate with the prior to get a shrunken gap and a
credible interval.

```{r stage3-posterior}
post <- get_posterior(fit)
head(as.data.frame(post))
```

The columns to compare are `estimate` (the raw gap) and `posterior_mean` (the
shrunken gap). The distance between them is empirical Bayes at work: a firm
measured imprecisely (large `se`) is pulled harder toward the prior centre,
because its own estimate carries less information; a precisely measured firm
barely moves. The `lower`/`upper` columns are the credible interval the report
card will display as-is.

## 6. Stage 4 — pairwise outranking (Pi)

The fourth stage turns 97 posteriors into a social-choice object. For every
ordered pair of firms $(i, j)$ it computes $\Pi_{ij} = \Pr(\theta_i > \theta_j)$,
the posterior probability that firm $i$'s true gap exceeds firm $j$'s. The result
is a $J \times J$ matrix:

```{r stage4-pairwise}
pw <- get_pairwise(fit)
pw$matrix[1:5, 1:5]
```

An entry near $1$ says the posterior strongly orders that row's firm above that
column's firm; an entry near $0.5$ says the data cannot separate them. The
companion `pw$ids` holds the firm identifiers in matrix order. The standalone
recomputation from a posterior is a fast one-liner, should you want it:

```{r stage4-standalone, eval = FALSE}
gp_pairwise(get_posterior(fit))   # recompute Pi from the posterior (fast)
```

The matrix `Pi` is the only thing the grade integer program reads. It carries no
ranking-superiority statement of any kind — only pairwise posterior order.

## 7. Stage 5 — the grade integer program

The fifth stage is the one that solves. It sorts the firms into a small number of
contiguous integer grades $\{1, \dots, k\}$ by solving an integer program over
`Pi`, penalized by $\lambda$: larger $\lambda$ lowers the separation threshold
$1/(1+\lambda)$ and yields more, finer grades; smaller $\lambda$ raises the
threshold and pools units into fewer, coarser grades. The calls you would write are:

```{r stage5-solve, eval = FALSE}
# Solve the whole path over a lambda grid (the engine the pipeline uses):
path <- gp_grade_path(get_pairwise(fit),
                      lambda_grid = c(0.25, 0.5, 1), control = gp_control())

# Or solve a single penalty in isolation:
gp_grade(get_pairwise(fit), lambda = 0.25)
```

Both **solve** an integer program over 97 firms (minutes with Gurobi), so we do
not run them here. We read the already-solved assignment off the loaded fit
instead:

```{r stage5-grades}
table(get_grades(fit))            # the grade composition: 2 / 81 / 14
fit$selected_grade$lambda         # the selected penalty
```

The `fit$selected_grade` slot is the assignment chosen at the selected penalty,
the `gp_grade_fit` at `lambda = 0.25`. Two firms land in the most-extreme grade,
81 in the middle, 14 in the least-extreme — the published `2 / 81 / 14`. Grade 1
is simply the most-extreme $\theta$ block; the label is an integer, not a verdict.

## 8. Stage 6 — the information-reliability frontier

The sixth stage reports the *cost of compression*. Each penalty $\lambda$ on the
grid yields one grade count, and with it a discordance rate (the share of
comparable pairs the grades run against) and `tau_bar` (the average between-grade
separation). The frontier table — one row per penalty — is cached on the solved
path, so reading it is instant:

```{r stage6-frontier}
s <- fit$grade_path$summary
sel <- s[abs(s$lambda - 0.25) < 1e-8, ]
sel[, c("lambda", "grade_count", "discordance_rate", "reliability", "tau_bar")]
```

At `lambda = 0.25` reliability is $1 - \text{DR} \approx 0.96$ and `tau-bar`
$\approx 0.21$: the three grades agree with the underlying posterior ordering on
about 96% of comparable pairs. Moving along $\lambda$ trades grade count against
reliability — that trade-off *is* the frontier.

To plot it, assemble a `gp_frontier` object; it is a fast read that does not
solve. Do **not** call `as.data.frame()` on it — read `fit$grade_path$summary` for
the numbers, as above, and use the object only to plot:

```{r stage6-figure, fig.width = 7, fig.height = 5, fig.cap = "The information-reliability frontier for the RACE fit (KRW Figure 6b): each point is one penalty on the lambda grid, plotted by its average between-grade separation against its reliability (1 - discordance rate). The selected lambda = 0.25 grading sits on this frontier.", fig.alt = "A scatter of points on a two-axis plot, one point per lambda penalty on the grid. The horizontal axis is reliability (one minus the discordance rate) and the vertical axis is the average between-grade separation. The points trace the trade-off between using fewer grades and keeping the grading consistent with the posterior ordering; the selected penalty lies on the frontier."}
fr <- gp_frontier(fit$grade_path, pairwise = get_pairwise(fit))
gp_plot_frontier(fr)
```

## 9. Stage 7 — the report card

The final stage assembles everything into the per-firm card: each firm with its
grade label, its posterior summary, its credible interval, and its original
estimate, sorted from most-extreme to least-extreme. `get_report_card(fit)`
returns that assembled card; for your own joins and plots, `as.data.frame(fit)`
gives the same content as a tidy 97 × 10 table, grade 1 first:

```{r stage7-table}
head(as.data.frame(fit), 8)
```

The columns are `id`, `label`, `grade`, `sort_rank`, `selected_lambda`,
`posterior_mean`, `lower`, `upper`, `estimate`, and `se`. The standalone assembler
builds the same card from the stage slots, and like the reads above it does
**not** solve:

```{r stage7-assemble, eval = FALSE}
gp_report_card(
  estimates      = fit$estimates,
  posterior      = get_posterior(fit),
  selected_grade = fit$selected_grade,
  grade_path     = fit$grade_path)
```

The signature figure is the card itself (KRW Figure 7) — each firm's posterior
gap with its interval, ranked and coloured by grade:

```{r stage7-figure, fig.width = 7, fig.height = 8, fig.cap = "The RACE report card (KRW Figure 7): each of the 97 firms shown as its posterior White-minus-Black callback gap (point) with its credible interval (horizontal line), one firm per row, ranked with the most-extreme firm on top and coloured by its grade (2 / 81 / 14).", fig.alt = "A horizontal caterpillar plot of 97 firms, one per row, ranked from most extreme at the top to least extreme at the bottom. Each row shows a point for the firm's posterior gap and a horizontal line for its credible interval. Points are coloured by grade into three groups: a small most-extreme grade at the top, a large middle grade, and a least-extreme grade at the bottom, in the 2 / 81 / 14 split."}
gp_plot_report_card(get_report_card(fit))
```

## 10. All in one

The seven stages of sections 3–9 are exactly what `krw_report_card()` chains in a
single call. Section 1's one-liner *is* this whole walk, run end to end:

```{r all-in-one, eval = FALSE}
fit <- krw_report_card(
  race, demographic = "race",
  control = gp_control(backend = "gurobi", precision_rule = "krw_gmm",
                       lambda_grid = c(0.25, 0.5, 1)))   # the grid must include 1
```

That call performs precision standardization → prior deconvolution → posterior
shrinkage → the pairwise matrix → the grade integer program → the frontier → the
report card, and returns the same `gp_fit` we loaded in section 1. Use the
one-call form in practice; reach for the stage accessors of this vignette when you
want to inspect or export an intermediate object.

## 11. The same workflow for gender

Nothing about the seven stages is specific to race. Swap the demographic and the
identical pipeline runs on the gender callback gap. The bundled gender fit reads
exactly like the race fit:

```{r gender-fit}
gfit <- gp_parity_fit("gender")
round(get_prior(gfit)$metadata$beta, 4)   # the gender precision exponent
table(get_grades(gfit))                    # the gender grade composition
```

The gender fit sorts the firms into four grades (`1 / 3 / 89 / 4`) with
$\beta = 1.2554$. The welfare reading of grade 1 follows the application, so the
labels are never hard-coded to one interpretation; every stage accessor of
sections 3–9 — `get_prior()`, `get_posterior()`, `get_pairwise()`,
`get_grades()`, `get_report_card()` — works identically on `gfit`.

## 12. Where to next

- **Reading and exporting** — `r vlink("a3-reading-and-exporting-results", "A3")`
  goes deeper on the accessors, the tidy table, and CSV / round-trip export.
- **Figures** — `r vlink("a4-figures", "A4")` is the full cookbook for all four
  KRW signature figures (frontier, posterior contrast, report card, discordance).
- **Solvers and calibration** — `r vlink("a5-solvers-and-calibration", "A5")`
  covers the open-source backends and the calibration harness.
- **Foundations** — `vignette("m1-foundations-and-notation")` formalizes the
  estimate → rank → grade pipeline and the notation this vignette kept light.
- **Method track, stage by stage** — `vignette("m2-precision-and-standardization")`
  derives the standardization of section 3;
  `vignette("m3-deconvolution-and-posterior")` derives the deconvolution and
  shrinkage of sections 4–5; `vignette("m4-grading-frontier-and-report-cards")`
  derives the integer program, frontier, and report card of sections 6–9; and
  `vignette("m5-two-level-and-calibration")` covers the two-level extension and
  calibration.

A reminder on reading the grades honestly: a grade is a sorting of firms by
posterior evidence, never a contest. The one-level race and gender core is
accepted within its declared scope, and the numbers here match the paper
($\beta = 0.5095$ vs the reported `0.510`; $\lambda = 0.25$ is the KRW baseline).

### Provenance

```{r provenance}
sessionInfo()
```

## References
