---
title: "A3: Reading and exporting results"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  A fitted gp_fit is a rich object: it carries the grades, the deconvolved
  prior, the per-firm posterior, the pairwise outranking matrix, the assembled
  report card, and the run-control that produced them. This vignette is the
  reference for getting all of that out and into your own workflow. We read the
  fit several ways for the console and the page (the result-first print box, the
  fuller summary with its Interpretation and Glossary, the tidy ten-column data
  frame, the six get_* accessors, and coef for a quick firm-to-grade lookup),
  then export it three ways (a CSV you drop into a paper table, an RDS that
  round-trips a gp_fit exactly, and an ASCII card you embed in a log via the
  format_*_cli renderers). Everything runs on the bundled RACE fit, and nothing
  here re-solves: every read is a fast lookup off a loaded object.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A3: Reading and exporting results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. What a fit holds

By the end of `r vlink("a2-the-grading-workflow", "A2")` you have a `gp_fit`: a
single object that bundles everything the KRW pipeline produced
[@kline2024discrimination] — the deconvolved prior, the per-firm posterior, the
$J \times J$ pairwise outranking matrix, the selected grading, the assembled
report card, and the run-control that drove the solve. This vignette is the
reference for reading that object and exporting it, with each read pointed at a
concrete downstream use: joining the card to firm covariates, dropping a block
into a paper table, handing off a reproducible artifact, or pasting a
result-first card into a build log.

We load the proven RACE fit **once** and read everything below off it. Loading is
instant because the fit ships pre-solved; **nothing in this vignette re-solves**.
Every read is a fast lookup off materialized slots — the accessors never call the
solver and recompute nothing.

```{r load-fit}
fit <- gp_parity_fit("race")
```

For the record, the one call that *produces* a fit in the first place is
`krw_report_card()` — it solves the integer program and takes minutes, so it is
shown here only as a pointer and never run:

```{r how-produced, eval = FALSE}
fit <- krw_report_card(
  race, demographic = "race",
  control = gp_control(backend = "gurobi", precision_rule = "krw_gmm",
                       lambda_grid = c(0.25, 0.5, 1)))
```

## 2. The quick read: `print()` / the console box

The fastest read is to print the fit. `print()` (which you get by typing `fit`
alone) emits a result-first ASCII box — it leads with the answer, not the inputs:

```{r print-fit}
fit
```

Line by line, this is the headline you would quote in a slide or a commit
message:

- The header — `gp_fit . 97 units . grades: 3 (2/81/14) . selected lambda = 0.25`:
  97 firms sorted into three grades, with 2 in the most-extreme grade, 81 in the
  middle, and 14 in the least-extreme grade, at the KRW baseline $\lambda = 0.25$.
- `reliability: (1 - DR)` — `0.96`: the share of comparable firm pairs on which
  the grades agree with the underlying posterior ordering. Higher is cleaner.
- `tau-bar` — `0.21`: the posterior expected Kendall rank agreement the grading
  carries across grade boundaries (between-grade blocks), in [-1, 1]; higher
  means the grades preserve more rank information.
- `backend` — `gurobi`: the solver that produced the grading.

A footer line points you onward — `summary()` for the full read,
`get_report_card()` for the ranked table. Grade 1 is simply the most-extreme
$\theta$ block under the data, not a verdict.

## 3. The full read: `summary()`

When you are writing up the result rather than glancing at it, `summary()` is the
read you want. It keeps every number from the box and adds three plain-English
blocks; it returns a typed `gp_fit_summary`, and printing it shows the full read:

```{r summary-fit}
summary(fit)
```

The three additions:

- **Interpretation** — the headline in words you can lift into a methods
  paragraph: the grades agree with the pairwise (posterior) ordering on about 96%
  of comparable pairs, and grade 1 is the most-extreme grade, never a contest.
- **Glossary** — each term defined in place: `reliability` $(1 - \text{DR})$,
  `tau_bar`, and `selected_lambda`, so a reader who has never opened the package
  can follow the output.
- **Provenance** — the backend (`gurobi`) and the selection rule
  (`baseline_lambda_0.25`), exactly what you cite to make the run reproducible.

Use `print()` for a glance and `summary()` when you need the reading and the
provenance on the page.

## 4. The tidy table: `as.data.frame(fit)`

For anything analytical — a join to firm covariates, a regression of grade on
firm size, a plot in your own house style — coerce the fit to a data frame.
`as.data.frame(fit)` returns the report-card rows as a tidy `97 × 10` table,
ranked with the most-extreme firm (grade 1) on top and row names reset:

```{r as-dataframe}
head(as.data.frame(fit), 8)
```

The ten columns are the working surface for downstream use:

- `id`, `label` — the firm identifier and its display label, your **join keys**
  against any external firm table.
- `grade` — the integer grade in $\{1, \dots, n\}$ (grade 1 is most extreme).
- `sort_rank` — the rank used to order the rows; order on it, facet on `grade`.
- `selected_lambda` — the penalty that produced this grading (`0.25`), carried on
  every row so the table is self-documenting.
- `posterior_mean`, `lower`, `upper` — the shrunken gap and its credible
  interval, reported as-is and never recomputed downstream.
- `estimate`, `se` — the raw gap and its standard error. Compare `estimate` with
  `posterior_mean` to watch empirical Bayes work: the noisier a firm's estimate,
  the harder its posterior is pulled toward the centre.

Because it is an ordinary data frame, this is the object you `merge()`,
`dplyr::left_join()`, or hand to `knitr::kable()` for a paper table. The
report-card object coerces to the identical ten columns, so you can start from
either end:

```{r card-dataframe}
identical(as.data.frame(get_report_card(fit)), as.data.frame(fit))
```

## 5. The accessors: the six `get_*` functions

A `gp_fit` is read through its accessor family, never by reaching into `$` slots.
Each accessor is a read-only lookup that recomputes nothing — reach for the one
whose output your downstream step needs. There are six.

**`get_grades(fit)`** — the selected grading as a named integer vector of length
97, names equal to the firm ids. This is what you `cbind()` onto a firm table or
pass to `table()` for the composition:

```{r get-grades}
head(get_grades(fit))
table(get_grades(fit))      # the grade composition: 2 / 81 / 14
```

**`get_report_card(fit)`** — the assembled `gp_report_card` object, the source of
the tidy table in section 4. Reach for it when you want the object that prints and
plots, not just the rows:

```{r get-report-card}
get_report_card(fit)
```

**`get_posterior(fit)`** — the native `gp_posterior` (per-firm means, standard
deviations, and intervals). Its data frame carries `estimate`, `se`, `id`,
`label`, `posterior_mean`, `posterior_sd`, `lower`, `upper`, `scale`, and more;
`estimate` versus `posterior_mean` is the shrinkage:

```{r get-posterior}
head(as.data.frame(get_posterior(fit))[, c("id", "estimate", "posterior_mean", "posterior_sd")])
```

**`get_prior(fit)`** — the deconvolved native `gp_prior`, a list carrying
`support`, `density`, `mean`, `scale`, `diagnostics`, and `metadata`. The
precision exponent $\beta$ — the number you cite to show the replication lands on
the paper — lives in the metadata:

```{r get-prior}
names(get_prior(fit))
round(get_prior(fit)$metadata$beta, 4)   # the precision exponent beta
```

gradepath lands on $\beta = 0.5095$, matching the `0.510` KRW report.

**`get_pairwise(fit)`** — the `gp_pairwise` social-choice input the grading is
solved over: the $J \times J$ matrix $\Pi_{ij} = \Pr(\theta_i > \theta_j)$ in
`$matrix`, with the firm order in `$ids`:

```{r get-pairwise}
pw <- get_pairwise(fit)
pw$matrix[1:5, 1:5]         # a corner of the J x J outranking matrix
head(pw$ids)
```

An entry near $1$ says the posterior strongly orders that row's firm above the
column's; an entry near $0.5$ says the data cannot separate them. The matrix
encodes pairwise posterior order only and carries no superiority statement.

**`get_control(fit)`** — the exact `gp_control()` that produced the fit, your
reproducibility receipt: the grid, the backend, the precision rule, the interval
level, the solver options, the seed, the schema version, and the provenance:

```{r get-control}
ctrl <- get_control(fit)
names(ctrl)
c(backend = ctrl$backend, precision_rule = ctrl$precision_rule)
```

## 6. `coef()`: a firm → grade lookup

`coef()` is the base generic an R user reaches for first, so a `gp_fit` answers it
with the grading: `coef(fit)` returns the same named integer vector as
`get_grades(fit)` — names = firm ids, values = the assigned grade — surfaced under
the name your muscle memory already knows:

```{r coef}
identical(coef(fit), get_grades(fit))
head(coef(fit))                 # e.g. krw_race_001 = 2
coef(fit)["krw_race_001"]       # the grade of one firm, by id
```

Because it is named, it doubles as a quick firm-to-grade lookup table: index it by
id to read any one firm's grade, or `data.frame(id = names(coef(fit)), grade =
coef(fit))` to spread the whole vector into two columns for a join. When you simply
need "which firm got which grade" without the posterior summaries, this is the
lightest read in the package; a grade label carries no ranking-superiority
statement.

## 7. The CLI renderers: embed a card in a log

`print()` and `summary()` write to the console, but sometimes you want the same
result-first card as a *character vector* you can splice into a build log, a
plain-text report, or an HTML `<pre>` block. The `format_*_cli` renderers are
exactly that: the formatter behind the console output, returned as one string per
line.

`format_gp_fit_cli(fit)` returns the box of section 2 as lines you `cat()` or
`writeLines()`:

```{r format-fit-cli}
writeLines(format_gp_fit_cli(fit))
```

`format_gp_report_card_cli()` does the same for the ranked card: it leads with the
unit count, grade composition, and selected $\lambda$, shows a HEAD of the ranked
table, and closes with the welfare footnote:

```{r format-card-cli}
writeLines(format_gp_report_card_cli(get_report_card(fit)))
```

Because each returns plain ASCII as a character vector, both compose with
everything that takes text — dropping a result-first card into a pipeline log is a
one-liner, `message(paste(format_gp_fit_cli(fit), collapse = "\n"))`, so a
teammate reading the log sees the grading without opening R. The renderer's
welfare note describes grade 1 as the most-extreme $\theta$ and is not hard-coded
to one reading: it adapts between the firms application (most discriminatory) and
the names application (best-treated), so the same code is honest for either.

## 8. Exporting

Three exports cover almost every hand-off.

**CSV.** The first is the analytical artifact: write the tidy table to CSV and it
lands in a spreadsheet, a paper table, or a co-author's pipeline. This is the
command you would run on your own machine (we do not write into the package
working directory here):

```{r export-csv, eval = FALSE}
write.csv(as.data.frame(fit), "report_card_race.csv", row.names = FALSE)
```

**RDS round-trip.** The second is the reproducible artifact: the whole `gp_fit`,
saved with `saveRDS()`, round-trips **exactly** — every slot, the prior, the
posterior, the pairwise matrix, and the control come back identical. We
demonstrate it live through a `tempfile()`, so the chunk both runs and proves the
integrity:

```{r export-rds}
path <- tempfile(fileext = ".rds")
saveRDS(fit, path)
identical(fit, readRDS(path))   # a gp_fit round-trips through RDS exactly
```

A `TRUE` here is the guarantee you rely on when you ship a solved fit alongside a
paper: a collaborator who `readRDS()`s the file recovers the identical object you
solved, with no re-solve and no drift. (This is precisely how the bundled parity
fits travel inside the package.)

**ASCII card to a file.** The third is the human-readable artifact: the
result-first card written to a text file, ready to attach to an issue or paste
into a README. Like the CSV, this is the command you would run yourself:

```{r export-card-txt, eval = FALSE}
writeLines(format_gp_report_card_cli(get_report_card(fit)), "card.txt")
```

## 9. The same reads for gender

Every read and export above is demographic-agnostic — the accessors, `coef()`, the
renderers, and the export idioms do not know or care which gap they describe. The
bundled gender fit answers every call of sections 2–8 identically, only with its
own numbers:

```{r gender}
gfit <- gp_parity_fit("gender")
table(get_grades(gfit))                     # the gender composition: 1 / 3 / 89 / 4
round(get_prior(gfit)$metadata$beta, 4)     # the gender precision exponent (1.2554)
head(coef(gfit))
```

The gender fit sorts the 97 firms into four grades (`1 / 3 / 89 / 4`), and its
precision exponent is $\beta = 1.2554$ — a different number from the race fit's
$0.5095$, as it should be, since the two demographics carry different gap series.
Its `summary()`, `as.data.frame(gfit)`, the six `get_*` accessors, the
`format_*_cli` renderers, and the CSV / RDS export all behave exactly as above:
`as.data.frame(gfit)` is the same ten columns, `coef(gfit)` is the same named
lookup, and `saveRDS(gfit, ...)` round-trips with the same `identical()`
guarantee. The welfare reading of grade 1 follows the application and is never
hard-coded to one interpretation.

## 10. Where to next

- **Getting started** — `r vlink("a1-getting-started", "A1")` is the gentle first
  contact, if you arrived here without a fit in hand;
  `r vlink("a2-the-grading-workflow", "A2")` walks the seven stages that fill the
  slots this vignette reads.
- **Figures** — `r vlink("a4-figures", "A4")` is the cookbook for all four KRW
  signature figures (frontier, posterior contrast, report card, discordance),
  built from these same accessors.
- **Solvers and calibration** — `r vlink("a5-solvers-and-calibration", "A5")`
  covers the open-source backends and the calibration harness.
- **Foundations** — `vignette("m1-foundations-and-notation")` formalizes the
  estimate → rank → grade pipeline, and the method track
  (`vignette("m2-precision-and-standardization")`,
  `vignette("m3-deconvolution-and-posterior")`,
  `vignette("m4-grading-frontier-and-report-cards")`,
  `vignette("m5-two-level-and-calibration")`) derives each stage whose output you
  read here.

A closing reminder on reading the output honestly: a grade is a sorting of firms
by posterior evidence, never a contest, and grade 1 is the most-extreme grade
rather than a judgment. The one-level race and gender core is accepted within its
declared scope, and the numbers here match the paper ($\beta = 0.5095$ versus the
reported `0.510`; $\lambda = 0.25$ is the KRW baseline).

### Provenance

```{r provenance}
sessionInfo()
```

## References
