---
title: "A4: The figure cookbook"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  gradepath ships four signature figures from Kline, Rose & Walters (2024) as
  ggplot2 verbs: the report card, the posterior-contrast heatmap, the
  information-reliability frontier, and the conditional discordance matrix. This
  cookbook builds each one off the bundled RACE fit in a single uniform recipe —
  purpose, lead chunk, default plot, one variant — and teaches you how to read
  it: what each axis, colour, and position means, and what the eye should catch
  first. It then covers the package theme, the ordinal grade palette, the
  autoplot() generic, and saving with ggsave(). No equations; pure visual
  literacy. Every figure is built off a loaded fit, so nothing here re-solves.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A4: The figure cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. Four figures, one recipe

gradepath turns a finished grading into four pictures, the same four that carry
the argument in Kline, Rose & Walters [@kline2024discrimination]. Each is a
single verb that returns a ggplot, and each answers one visual question:

- **The report card** — where does every firm sit, and how sure are we?
- **The posterior-contrast heatmap** — how cleanly do the grade blocks separate?
- **The information-reliability frontier** — what does compressing into fewer
  grades cost, and how do simpler scoring rules compare?
- **The conditional discordance matrix** — between which grades are the cuts
  cleanest?

This cookbook gives each figure the same four-part recipe so you can navigate by
what you need to communicate:

1. **Purpose** — the one visual question it answers.
2. **Lead chunk** — the few lines that prepare the object to plot.
3. **Default plot** — the figure as shipped, with a reading guide.
4. **One variant** — a single argument worth knowing.

Two things hold throughout. First, every `gp_plot_*()` verb *returns* a ggplot
and renders nothing on its own — you print it (which a code chunk does
automatically), and you can keep extending it with `+ theme_*()` or `+ scale_*`.
Second, nothing here solves: we load the proven RACE fit once and build all four
figures off it.

```{r load-fit}
fit <- gp_parity_fit("race")
fit
```

That is the bundled 97-firm RACE report card — grades `2 / 81 / 14` at the KRW
baseline `lambda = 0.25`, reliability `0.96`, `tau-bar = 0.21`. Everything below
reads off `fit`. A reminder before we start drawing: the grades are a *sorting by
posterior evidence*, never a contest, and grade 1 is simply the most-extreme
block — the figures show structure, not a ranking verdict. The one-level race
core is accepted within its declared scope, and these figures are descriptive.

## 2. Figure 1 — the report card

**Purpose.** Show every firm at once: its shrunken estimate, its uncertainty, and
the grade it landed in. This is KRW Figure 7, the headline figure.

**Lead.** Pull the assembled card off the fit with `get_report_card()` and hand
it to the verb.

```{r lead-report-card}
card <- get_report_card(fit)
```

**Default plot.** A horizontal point-and-interval ("caterpillar"): one firm per
row, the most-extreme firm on top.

```{r figure-report-card, fig.width = 7, fig.height = 8, fig.cap = "The RACE report card (KRW Figure 7): each of the 97 firms shown as its posterior White-minus-Black callback gap (point) with its credible interval (horizontal line), one firm per row, ranked with the most-extreme firm on top and coloured by its grade in the 2 / 81 / 14 split.", fig.alt = "A horizontal caterpillar plot of 97 firms, one per row, ranked from most extreme at the top to least extreme at the bottom. Each row shows a point for the firm's posterior gap and a horizontal line for its credible interval. Points are coloured by grade into three groups: a small most-extreme grade at the top, a large middle grade, and a least-extreme grade at the bottom."}
gp_plot_report_card(card)
```

**How to read it.** Three things carry the meaning, and reading them in order is
the whole skill:

- **Vertical position is rank.** Top row is the most-extreme firm; the eye should
  travel top-to-bottom as "most to least extreme."
- **Horizontal position is the posterior gap.** Where the point sits on the
  x-axis is the firm's shrunken estimate — already pulled toward the centre by
  empirical Bayes, so it is the honest gap, not the raw one.
- **Line length is uncertainty.** A long horizontal line is an imprecisely
  measured firm; a short one is a tightly measured firm. The first thing the eye
  should catch is *which intervals overlap*: where two firms' lines cover the
  same ground, the data cannot separate them, and the colour blocks respect
  exactly that.

The **colour** is the grade. The three colour bands stack top-to-bottom, so the
grade boundaries read off as the points where the colour changes — and they fall
where the intervals stop overlapping, not at arbitrary cut-points.

**Variant — `max_rows`.** On 97 rows the labels crowd; show only the top of the
ranking when the extremes are the story. The verb adds a "Showing N of M"
subtitle so the truncation is never silent.

```{r figure-report-card-top, fig.width = 7, fig.height = 5.5, fig.cap = "The same RACE report card restricted to the 20 most-extreme firms (max_rows = 20): the top of the ranking at full height, with a subtitle noting that 20 of 97 firms are shown.", fig.alt = "A horizontal caterpillar plot showing only the top 20 firms of the report card, one per row, each with a point for its posterior gap and a horizontal line for its credible interval, coloured by grade, with a subtitle indicating that 20 of 97 firms are displayed."}
gp_plot_report_card(card, max_rows = 20)
```

Other one-argument turns: `order = "grade"` groups the rows by grade block
instead of by rank, and `zero_line = "on"` always draws the dashed reference at
zero (the default `"auto"` draws it only when intervals straddle zero, so you can
see at a glance which firms are distinguishable from no gap at all).

## 3. Figure 2 — the posterior-contrast heatmap

**Purpose.** Look at the *structure* behind the grades: the full matrix of
pairwise posterior contrasts, reordered so the grade blocks sit together. This is
KRW Figure 5a. Where Figure 1 shows firms one at a time, this shows every firm
against every other at once.

**Default plot.** The verb reads the pairwise matrix and the grades straight off
the fit, so there is no lead chunk — pass it `fit`.

```{r figure-posterior-contrast, fig.width = 6.5, fig.height = 6, fig.cap = "The grade-sorted posterior-contrast heatmap for the RACE fit (KRW Figure 5a): the 97 by 97 pairwise outranking matrix with rows and columns reordered by grade so the grade blocks are contiguous (grade 1 at the top-left), black segments separating the blocks, and cells reaching the different-grade threshold outlined.", fig.alt = "A square heatmap of the 97-by-97 pairwise posterior-contrast matrix. Rows and columns are sorted by grade so the grade blocks form contiguous squares along the diagonal, with the most-extreme grade at the top-left. Cell shade encodes the posterior probability that the row firm is more extreme than the column firm; black segments separate the grade blocks and threshold-reaching cells are outlined in red."}
gp_plot_posterior_contrast(fit)
```

**How to read it.** Read it as a grid of light and dark, block by block:

- **Each cell is one ordered pair.** Its shade is the posterior probability that
  the row firm is more extreme than the column firm — dark means "almost surely
  ordered this way," and a washed-out middle grey means "the data cannot tell
  these two apart."
- **The diagonal runs corner to corner**, and rows/columns are sorted by grade
  with **grade 1 in the top-left**. So the picture should read as a gradient:
  dark in the upper-right triangle (more-extreme firms over less-extreme ones),
  light in the lower-left.
- **The black segments mark the grade boundaries.** The eye should catch the
  *square blocks* they carve out along the diagonal: a clean grading shows
  near-uniform shade *inside* each block (firms in the same grade are hard to
  separate) and sharp contrast *across* blocks (different grades really are
  different). The outlined cells are the pairs that reach the different-grade
  threshold — the evidence that earned a cut.

This is a structure plot, not a contest: a dark cell says the posterior orders
one firm above another, nothing more.

**Variant — `threshold = FALSE`.** Drop the threshold outlines to see the raw
contrast surface with no annotation — useful when you want the shading alone to
tell the story.

```{r figure-posterior-contrast-raw, fig.width = 6.5, fig.height = 6, fig.cap = "The same posterior-contrast heatmap with threshold = FALSE: the grade-sorted outranking matrix shown as raw shading, without the red outlines on cells that reach the different-grade threshold.", fig.alt = "The same grade-sorted square heatmap of pairwise posterior contrasts, but without the threshold outlines. Only the cell shading and the black grade-block separators remain, so the block structure reads off the shading alone."}
gp_plot_posterior_contrast(fit, threshold = FALSE)
```

## 4. Figure 3 — the information-reliability frontier

**Purpose.** Show the trade-off the frontier makes explicit: between
**reliability** (`1 - DR`, the x-axis — how often the grades agree with the
pairwise posterior ranking) and the **information** the grades carry (`tau-bar`,
the y-axis — the posterior expected rank agreement). A finer grading (more
grades) carries more information but is less reliable; pooling into fewer grades
raises reliability at the cost of information. The frontier traces that trade-off
across the grade path and overlays simpler scoring rules, so you can see where
the gradepath grading sits relative to them. This is KRW Figure 6b.

**Lead.** The frontier is its own object. Build it from the fit's grade path and
pairwise, and pass `benchmark_scores` — one value per firm, named by firm id — to
overlay the naive rules. Here we overlay the raw gap, the full posterior mean,
and a half-shrinkage rule.

```{r lead-frontier}
post <- get_posterior(fit)
ids  <- get_pairwise(fit)$ids
fr <- gp_frontier(
  fit$grade_path, pairwise = get_pairwise(fit),
  benchmark_scores = list(
    raw_estimate     = setNames(post$estimate,       ids),   # the raw gap
    posterior_mean   = setNames(post$posterior_mean, ids),   # full shrinkage
    linear_shrinkage = setNames(0.5 * post$estimate, ids)))  # half shrinkage
```

**Default plot.** Plot `fr`. This figure carries a two-row legend at the bottom,
so give it a slightly **wider, shorter canvas** (`fig.width = 8`,
`fig.height = 5.5`) so the legend is not clipped.

```{r figure-frontier, fig.width = 8, fig.height = 5.5, fig.cap = "The information-reliability frontier for the RACE fit (KRW Figure 6b): the optimal-grade path traced in reliability (1 - discordance rate, x-axis) against the posterior expected agreement tau-bar (y-axis), with the selected lambda = 0.25 grading marked and the three naive scoring rules overlaid as dominated points.", fig.alt = "A scatter-and-line plot. The horizontal axis is reliability, one minus the discordance rate; the vertical axis is the posterior expected agreement tau-bar. A connected path of points traces the grade frontier from few grades to many; the selected grading is highlighted, and three naive scoring rules appear as separate points sitting below and inside the frontier. Colour and shape are merged into a single two-row legend at the bottom."}
gp_plot_frontier(fr)
```

**How to read it.** The two axes are the whole idea, and the benchmarks are the
payoff:

- **The x-axis is reliability, `1 - DR`** — how often the grades agree with the
  underlying posterior ordering. Further right is cleaner.
- **The y-axis is `tau-bar`**, the posterior expected agreement carried by the
  grading. Higher is more informative.
- **The connected points are the frontier itself.** Walk it from one end to the
  other and you are trading grade count against reliability; the **highlighted
  point** is the selected `lambda = 0.25` grading, and the subtitle reports its
  exact `(1 - DR, tau-bar)` coordinates.
- **The overlaid benchmark points are the simple scoring rules** — the raw gap,
  full shrinkage, half shrinkage. What the eye should catch is that they sit
  *below and inside* the frontier: the gradepath grading reaches reliability they
  cannot, which is the figure's point. Colour and shape are merged into one
  two-row legend at the bottom, and there are no in-plot text labels, so read
  each series off that single legend. The benchmarks are descriptive reference
  points, not opponents.

**Variant — `benchmarks = FALSE`.** Drop the naive rules to show the bare
frontier — the trade-off curve on its own, when the benchmarks are not the story.

```{r figure-frontier-bare, fig.width = 8, fig.height = 5.5, fig.cap = "The bare information-reliability frontier (benchmarks = FALSE): just the optimal-grade path in reliability against tau-bar, with the selected lambda highlighted and no naive-rule overlays.", fig.alt = "The same frontier plot with the benchmark points removed. Only the connected path of grade points remains, with the selected grading highlighted, plotted in reliability on the horizontal axis against tau-bar on the vertical axis."}
gp_plot_frontier(fr, benchmarks = FALSE)
```

The other small turns: `annotate = FALSE` drops the coordinate readout from the
subtitle for a cleaner caption, and `highlight = "all"` marks every path point
instead of only the selected one.

## 5. Figure 4 — the conditional discordance matrix

**Purpose.** Ask *between which grades the cuts are cleanest.* For each pair of
grades it reports the discordance rate — the posterior share of cross-grade firm
pairs whose grade ordering runs against the posterior ordering. This is KRW
Figure 10a, and it reads off the same `fr` object.

**Default plot.** A lower-triangular heatmap of the between-grade discordance
rates.

```{r figure-discordance, fig.width = 6, fig.height = 5, fig.cap = "The conditional discordance matrix for the RACE fit (KRW Figure 10a): a lower-triangular heatmap of the between-grade discordance rate for each pair of grades, with the more-extreme grade on the columns (grade 1 at the left) and the less-extreme grade on the rows, each cell annotated with its rate. Small off-diagonal values mean clean cuts between grades.", fig.alt = "A lower-triangular heatmap with one cell per pair of grades. Columns run from the most-extreme grade on the left; rows run from the second grade at the top down. Each cell is shaded by its between-grade discordance rate, darker for higher discordance, with the numeric rate written inside. The off-diagonal cells are light, indicating clean separation between grades."}
gp_plot_discordance(fr)
```

**How to read it.** It is a small triangle, so read every cell:

- **Each cell is one grade pair**, with the more-extreme grade on the **columns**
  (grade 1 at the left) and the less-extreme grade on the **rows**.
- **The cell value and its shade are the discordance rate** for that pair — the
  posterior share of cross-grade firm pairs ordered against the posterior.
  **Darker is more discordant.**
- **The eye should catch the off-diagonal cells being light.** A light
  off-diagonal means the cut between those two grades is clean: almost every
  cross-grade pair is ordered the way the posterior orders it. A dark cell would
  flag two adjacent grades the data struggle to keep apart. On the RACE fit the
  off-diagonal stays small — the grades separate cleanly.

**Variant — `annotate = FALSE`.** Drop the printed values for a shading-only
heatmap, cleaner when you have many grades or want the colour alone to read.

```{r figure-discordance-bare, fig.width = 6, fig.height = 5, fig.cap = "The same conditional discordance matrix with annotate = FALSE: the lower-triangular between-grade heatmap shown by shading only, without the numeric rate in each cell.", fig.alt = "The same lower-triangular discordance heatmap, but without numbers in the cells. Each grade-pair cell is shaded by its discordance rate, darker for higher discordance, so the between-grade separation reads off the shading alone."}
gp_plot_discordance(fr, annotate = FALSE)
```

## 6. The theme

Every gradepath verb already draws in the package house style,
`theme_gradepath()` — a clean theme built on `ggplot2::theme_minimal()` with the
top and right spines removed, a light horizontal grid, no vertical grid, and the
legend at the bottom. The setup chunk made it the session default with
`theme_set(theme_gradepath())`, so your *own* plots match the four figures above
without any extra call.

You apply it explicitly two ways: set it once as the default, or add it to a
single plot with `+`.

```{r theme-demo, fig.width = 7, fig.height = 4, fig.cap = "theme_gradepath() applied to a plain ggplot of each firm's raw gap against its posterior gap: the gradepath house style with bottom legend, light horizontal grid, and no top or right spine.", fig.alt = "A scatter plot of 97 firms, each point placed by its raw callback gap on the horizontal axis and its posterior gap on the vertical axis, drawn in the gradepath house style: minimal background, a light horizontal grid only, and no top or right axis spine."}
ggplot(as.data.frame(fit), aes(estimate, posterior_mean)) +
  geom_abline(linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.7) +
  labs(x = "raw gap", y = "posterior gap", title = "RACE — raw vs posterior") +
  theme_gradepath()
```

Because the verbs return ordinary ggplots, you can also re-apply or extend the
theme on any of the four figures — for example
`gp_plot_discordance(fr) + theme_gradepath(base_size = 13)` to bump the font for
a slide.

## 7. The grade palette

The four figures colour grades with one shared ordinal scale, a perceptually
uniform, colour-blind-safe viridis ramp in which **grade 1 (the most-extreme
block) is the dark, high-contrast end**. To colour your own grade-keyed plot the
same way, add `scale_colour_gradepath()` (spelled `scale_color_gradepath()` too)
for the `colour` aesthetic, or `scale_fill_gradepath()` for `fill`.

```{r palette-demo, fig.width = 7, fig.height = 4.5, fig.cap = "The ordinal grade palette applied by hand: the 97 firms plotted by posterior gap against rank and coloured by their grade with scale_colour_gradepath(), so a custom plot uses the exact grade colours of the report card.", fig.alt = "A scatter plot of 97 firms by rank on the horizontal axis and posterior gap on the vertical axis, with each point coloured by its grade using the gradepath ordinal palette: the most-extreme grade in the dark, high-contrast colour and the least-extreme grade in the light end of the viridis ramp."}
card <- as.data.frame(fit)
ggplot(card, aes(sort_rank, posterior_mean, colour = factor(grade))) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_gradepath(name = "grade") +
  labs(x = "rank", y = "posterior gap")
```

The point of the shared scale is consistency: a firm's grade reads the same
colour on your custom plot as on the report card, so a reader's eye carries the
mapping across figures. For a `fill` aesthetic (tiles, bars) use
`scale_fill_gradepath()`; pass `n_grades` to fix the palette length when the data
do not contain every level. The colour is an *ordinal* encoding of evidence,
never a value judgement.

## 8. autoplot()

If you would rather not remember which verb draws which figure, the ggplot2
`autoplot()` generic dispatches on the gradepath object — and returns the *same*
ggplot the matching `gp_plot_*()` verb would, without attaching ggplot2.
`autoplot()` on a frontier draws the frontier; on a report card it draws the
caterpillar.

```{r autoplot-frontier, fig.width = 8, fig.height = 5.5, fig.cap = "autoplot() on the gp_frontier object: the same information-reliability frontier as gp_plot_frontier(fr), dispatched through the ggplot2 generic.", fig.alt = "The information-reliability frontier produced by the autoplot generic on the frontier object: the connected grade path in reliability against tau-bar with the selected grading highlighted and the naive rules overlaid, identical to the gp_plot_frontier output, with a single two-row bottom legend."}
autoplot(fr)
```

```{r autoplot-report-card, fig.width = 7, fig.height = 8, fig.cap = "autoplot() on the report-card object: the same caterpillar as gp_plot_report_card(get_report_card(fit)), dispatched through the ggplot2 generic.", fig.alt = "The report-card caterpillar produced by the autoplot generic on the report-card object: 97 firms one per row, ranked, each with a point for its posterior gap and a line for its credible interval, coloured by grade, identical to the gp_plot_report_card output."}
autoplot(get_report_card(fit))
```

These are S3 methods, so call `autoplot(x)` (not `autoplot.gp_frontier(x)`) and
let dispatch pick the figure. `autoplot()` also works on a bare `fit` — it draws
the frontier — and forwards extra arguments to the underlying verb, so
`autoplot(get_report_card(fit), max_rows = 20)` behaves like the variant in
section 2.

## 9. Saving figures

Every verb returns a ggplot, so save any of the four with `ggsave()`. Pass the
plot object, a filename, and a size. A useful rule of thumb from the canvases
above: the report card wants a *tall* aspect (many rows), the heatmaps want a
*square* one, and the frontier wants a *wide* one (the two-row bottom legend).

```{r save-demo-eval, eval = FALSE}
# The report card: tall.
ggsave("fig_report_card.png", gp_plot_report_card(get_report_card(fit)),
       width = 7, height = 8, dpi = 150)
# The heatmaps: square.
ggsave("fig_posterior_contrast.png", gp_plot_posterior_contrast(fit),
       width = 6, height = 6, dpi = 150)
ggsave("fig_discordance.png", gp_plot_discordance(fr),
       width = 6, height = 5, dpi = 150)
# The frontier: wide, so the bottom legend is not clipped.
ggsave("fig_frontier.png", gp_plot_frontier(fr),
       width = 8, height = 5.5, dpi = 150)
```

To verify the recipe end-to-end without leaving a file in your working directory,
write to a temporary path and check it exists:

```{r save-demo-tempfile}
out <- file.path(tempdir(), "fig_frontier.png")
ggsave(out, gp_plot_frontier(fr), width = 8, height = 5.5, dpi = 150)
file.exists(out)
```

For a paper, swap the `tempdir()` path for your own filename and reach for a
vector device (`device = cairo_pdf` for a PDF) when the journal wants vector art.

## 10. Where to next

You now have all four KRW signature figures off a single loaded fit, plus the
theme, the palette, `autoplot()`, and saving. The four are descriptive summaries
of a finished grading: nothing in these figures asserts more than the posterior
evidence supports — the grades are a sorting, never a contest.

- **The workflow behind the fit** — `r vlink("a2-the-grading-workflow", "A2")`
  walks the seven stages that produce the object every figure here reads.
- **Getting started** — `r vlink("a1-getting-started", "A1")` is the gentle first
  contact, with the single most informative figure (the report card) in context.
- **Reading and exporting** — `r vlink("a3-reading-and-exporting-results", "A3")`
  covers the accessors, the tidy table, and CSV / round-trip export.
- **Solvers and calibration** — `r vlink("a5-solvers-and-calibration", "A5")`
  covers the open-source backends and the calibration harness.
- **The figures, formalized** —
  `r vlink("m4-grading-frontier-and-report-cards", "M4")` derives the frontier,
  the discordance matrix, and the report card with the math this cookbook
  deliberately skipped; `r vlink("m1-foundations-and-notation", "M1")` fixes the
  notation.

### Provenance

```{r provenance}
sessionInfo()
```

## References
