---
title: "A5: Solvers and calibration"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This is the practical-operations vignette — the last of the applied track,
  for running gradepath day to day rather than reading its output. It covers
  four things. First, choosing a grade integer-program backend: Gurobi is the
  recommended default and by far the fastest here, while the license-free HiGHS
  is the open fallback. Second, sizing a run before you commit to a minutes-long
  solve, with the solve-free gp_preview(). Third, the seeded Monte-Carlo
  gp_calibrate() harness — what it checks, the budget heads-up it prints, and its
  honest scope as an in-package diagnostic rather than a Table E1 reproduction.
  Fourth, reading the package's sharpened error messages so a mistake names
  exactly what went wrong. The package installs, loads, and runs its tests with
  no Gurobi license; only an actual grade solve needs one.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A5: Solvers and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

## 1. Running gradepath in practice

The earlier applied vignettes were about *output*: load a fit, read its grades,
draw its figures. This one is about *operations* — the four things you actually
manage when you run gradepath on your own input rather than off a bundled fit:

- **The solver.** The grade step is an integer program, and it needs a backend.
  Which one, and what does each cost you?
- **Sizing.** A real 97-firm solve takes minutes. How do you confirm a run is
  well-formed and learn how big it is *before* paying for it?
- **Calibration.** Does the grading method's frequentist behaviour hold up under
  simulation, and how far can you trust the check?
- **Errors.** When you assemble something by hand and get it wrong, does the
  package tell you exactly what is missing?

Nothing in this vignette solves a grade problem on render. The solve-free pieces
— `gp_preview()`, building a control, the sharpened-error demos — run live; the
solving pieces — `gp_calibrate()`, `krw_report_card()` — are shown but not
executed, exactly as the package documents. For the live demos we use a real
bundled fit and the bundled tiny raw input:

```{r load-inputs}
fit <- gp_parity_fit("race")
inp <- readRDS(system.file("extdata/examples/tiny_input.rds", package = "gradepath"))
```

`fit` is the proven 97-firm RACE report card — grades `2 / 81 / 14` at the KRW
baseline `lambda = 0.25` [@kline2024discrimination]; `inp` is a tiny raw input,
the per-unit `theta_hat` and `s` series, that we will preview without solving.
Throughout, the grades are a *sorting by posterior evidence*, never a contest:
grade 1 is simply the most-extreme block, and the one-level race and gender core
is accepted within its declared scope.

## 2. Choosing a solver

gradepath grades units by solving an integer program, and that program needs a
backend solver. The package follows Kline, Rose & Walters
[@kline2024discrimination] on the default, and offers an open fallback.

- **Gurobi — the recommended default.** It is the solver KRW used, and on this
  problem it is by far the fastest backend; a real 97-firm solve lands in a few
  minutes. The one cost is that Gurobi is commercial and needs a license (free
  academic licenses are available).
- **HiGHS — the open, license-free fallback** (`backend = "highs"`, via the
  `highs` package). It is several times slower on a real 97-firm solve, but it
  installs with no license and reaches the same grading. (Two more license-free
  routes, `"glpk"` and `"symphony"`, go through ROI plugins; HiGHS is the one to
  reach for first.)

You select the backend two ways. Per run, set it on the control object — this
just *builds a configuration* and does not solve, so it runs instantly:

```{r control-highs}
ctrl <- gp_control(backend = "highs")   # license-free; builds instantly (no solve)
ctrl$backend
```

Or set it once as the session-wide default. `gp_control()` with no backend
argument reads `getOption("gradepath.backend", "gurobi")`, so this option becomes
the default every later `gp_control()`, `gp_preview()`, and `krw_report_card()`
will use:

```{r option-global, eval = FALSE}
options(gradepath.backend = "gurobi")   # global default for the session
```

A control carries more than the backend — the `lambda_grid`, the
`precision_rule`, the credible-interval level, and solver knobs like `mip_gap`
and `time_limit` — but for solver choice the `backend` field is the one that
matters. You then hand the control to `krw_report_card()`, `gp_grade_path()`, or
`gp_preview()` via `control =`.

Two honest points, stated plainly. First, the backend is checked by the solver
*at call time*, not when you build the control — `gp_control(backend = "highs")`
succeeds whether or not the `highs` package is installed; you find out when you
actually solve. Second, and more reassuring: **the package installs, loads, and
runs its full structural test suite with no Gurobi license at all.** Only an
actual grade solve needs a backend — Gurobi or HiGHS. So you can explore the
bundled fits, preview runs, read tables, and draw figures license-free, and you
reach for a solver only when you grade fresh data. On attach, gradepath prints a
one-line note reporting the backend it resolved, so you always know which solver
a fresh solve *would* use. A HiGHS solve reaches the same grading Gurobi does; it
is simply slower, and a non-optimal solve is never relabelled optimal — the
solver status is reported honestly.

## 3. Sizing a run with `gp_preview()`

A real solve costs minutes, so before you commit you want two things:
confirmation that your input is well-formed, and an estimate of how much work the
run is. `gp_preview()` gives you both *without solving anything*.

It takes the same input and `demographic` you would hand to `krw_report_card()`,
validates the input grammar, and reports the **plan** — the unit ids, the
inferred workflow, the backend, the frontier penalty grid, and the number of
grade-IP solves the run would perform. It runs **no** beta-GMM, **no**
deconvolution, **no** posterior, **no** pairwise, and **no** solver stage. So it
is instant, needs no license, and we run it live:

```{r preview}
pv <- gp_preview(inp, demographic = "race")
pv                   # result-first print of the planned run
```

The print is result-first: it shows the planned run at a glance. The object is a
`gp_preview` you can read field by field — the operational ones are:

```{r preview-fields}
pv$n_units           # units the run would grade
pv$estimated_solves  # number of grade-IP solves (= length of the lambda grid)
pv$backend           # the backend the run would use
pv$workflow          # the inferred workflow ("one_level_independence" here)
```

The full slot set is all thirteen of `ids` (the canonical unit-id order parsed
from the input), `n_units` (the unit count, `length(ids)`), `demographic`,
`groups` (the parsed grouping, or `NULL`), `workflow`, `lambda_grid` (the penalty
grid the run would sweep), `estimated_solves`, `backend`, the `control` it was
built from, the `status` — a `gp_status` object that reads `OK` for a clean
one-level plan; and three bookkeeping slots: `schema_version` (the object's
schema tag, here `"v2"`), `provenance` (a short record of how the preview was
produced — `producer = "gp_preview"`, `no_solve = TRUE`), and `warnings` (any
diagnostics raised while building the plan — empty for a clean preview).

`estimated_solves` is the headline number: each entry in the lambda grid is one
integer-program solve, so on a real fit this is how many minutes-long solves the
run will perform. The default grid is dense (`seq(0, 1, by = 0.01)`, 101 points),
so a production run narrows it — for the parity result the grid is
`c(0.25, 1)`, two solves. This is exactly how you size a 97-firm run before
committing: preview it, read `n_units` and `estimated_solves`, multiply by the
per-solve cost your backend gives you (Gurobi minutes, HiGHS more), and decide
whether to run it now or overnight. Preview the tiny input first to learn the
shape; preview the real input to learn the bill.

## 4. Calibration with `gp_calibrate()`

`gp_calibrate()` is the package's seeded Monte-Carlo calibration harness
(GP-DEC-18-A). It asks a different question from everything
above: *does the grading method's own frequentist behaviour hold up under
simulation?* It treats a fitted prior as ground truth, simulates `n_sim`
synthetic datasets from it, re-runs the full one-level pipeline
(`krw_report_card()`) on each, and checks three guarantees *qualitatively*:

1. the empirical discordance rate (DR) tracks `dr_target`;
2. the posterior credible intervals cover the truth near `ci_level`;
3. the empirical grading's regret against an oracle that *knows* the true prior
   is small and **non-negative by construction** (both gradings are scored on the
   same truth, so regret cannot go below zero).

The signature is

```{r calibrate-sig, eval = FALSE}
gp_calibrate(fit, n_sim = 200L, seed = 15238L,
             dr_target = 0.05, ci_level = 0.9, min_ok = 1L)
```

and it returns a compact `gp_calibration` carrying `n_sim`, `seed`,
`characteristic` (`"race"` or `"gender"`), `n_ok` and `n_failed` (successful and
skipped draw counts, which sum to `n_sim`), `dr_mean` and `dr_target`, `coverage`
and `ci_level`, `regret_mean`, the two boolean verdicts `dr_ok` (DR within `0.01`
of target) and `coverage_ok` (coverage within `0.02` of `ci_level`), plus
provenance. The harness is robust to refit non-convergence: a draw whose refit
fails or degenerates is skipped and counted in `n_failed`, the means are taken
over the successful draws only, and if fewer than `min_ok` succeed it errors
rather than averaging noise.

**The call — shown, not run.** `gp_calibrate()` re-solves the grade integer
program *once per draw*, and the package states it **never runs on package load
or on vignette render**. So we show it only, with `eval = FALSE`; keep `n_sim`
small because each draw costs a solve:

```{r calibrate-call, eval = FALSE}
cal <- gp_calibrate(fit, n_sim = 5, seed = 15238)   # each draw re-solves the IP
cal                                                  # compact, result-first one-liner
summary(cal)                                         # one-row typed data frame
```

**The budget heads-up.** Because a run can be long, `gp_calibrate()` prints a
one-line, suppressible `message()` *before* the draw loop whenever
`n_sim * n_units` exceeds 600 — informational, never changing the result. On this
97-unit fit, `n_sim = 20` gives `20 * 97 = 1940 > 600`, so you would see:

> `gp_calibrate(): about to run 20 refit+grade solves on 97 units. The grade
> step is an integer program (often minutes per solve on a large fit), so this
> can take a long time. For a quick check use a smaller n_sim or a smaller,
> well-identified fit; gp_preview() gauges one solve's cost.`

Silence it with `suppressMessages()` once you have sized the run.

**The honest scope.** State this plainly:

- `gp_calibrate()` is a **qualitative, in-package diagnostic — not a reproduction
  of KRW Table E1.** There is no parseable E1 artifact and the Matlab
  Monte-Carlo stream cannot be reproduced, so an exact E1 match is out of reach.
  The `15238` default seed is a **cosmetic homage** to the reference seed, not
  Matlab parity (the provenance records `matlab_rng_parity = FALSE`); it only
  fixes in-package reproducibility — the same seed yields identical output, and
  your global RNG stream is restored afterward.
- It is **not a headline result and not part of any default scorecard.**
- On **tiny or weakly identified fits the verdicts are frequently `FALSE`** — the
  small-N beta-GMM is noisy and its r-scale can be ill-identified — so a
  meaningful calibration needs a **well-identified fit** (KRW's real multi-firm
  input). The harness reports honestly either way; it never tunes the data to
  force a pass.

For the full calibration scope, including how this sits relative to the deferred
E1 Monte-Carlo, see `r vlink("m5-two-level-and-calibration", "M5")`.

## 5. Troubleshooting — the sharpened errors

gradepath's errors are written to name *exactly* what went wrong and how to fix
it. Each demo below errors **immediately**, with no solve, so we run them live
through `try()` (which lets the chunk continue past the error).

**Stage-wise assembly with a missing argument.** If you assemble a report card
from individual stage objects and forget one, the error names every missing
argument and hands you a copy-paste recipe built from a fit's slots:

```{r err-assembly}
try(gp_report_card(grade_path = fit$grade_path))
```

The message names the missing `posterior` and `selected_grade` and gives the full
`gp_report_card(estimates = fit$estimates, posterior = get_posterior(fit),
selected_grade = fit$selected_grade, grade_path = fit$grade_path)` form — no
guessing which slot you dropped.

**A bad control value.** `gp_control()` validates each field as you build it, so a
typo errors clearly at construction time rather than surfacing later inside a
solve:

```{r err-control}
try(gp_control(precision_rule = "nope"))
```

**A grouped input is previewed, not run.** Grouped, two-level execution is the M2
surface, not the one-level M1 monolith. If you hand `gp_preview()` a non-`NULL`
`groups` vector, it does not error outright — it returns a `gp_preview` whose
`status` is `GROUPS_ERROR`, with an explanatory warning, so you can see the
planned (but unrun) grouped workflow:

```{r err-groups}
pv_grouped <- suppressWarnings(
  gp_preview(inp, demographic = "race",
             groups = rep(c("A", "B"), length.out = pv$n_units))
)
pv_grouped$status
pv_grouped$workflow
```

The `GROUPS_ERROR` status and the `"grouped_pending"` workflow are the signal:
the one-level pipeline will not run a grouped input, and the two-level surface
lives in the m-track — see `r vlink("m5-two-level-and-calibration", "M5")`. Treat
the preview as a routing check, not a failure.

## 6. Where to next

You now have the operational toolkit: pick a backend, size a run with
`gp_preview()`, calibrate within scope with `gp_calibrate()`, and read the
sharpened errors. The applied track ends here; the method track derives what
these operations sit on top of.

- **The applied track** —
  `r vlink("a1-getting-started", "A1")` is the gentle first contact,
  `r vlink("a2-the-grading-workflow", "A2")` walks the seven stages behind a fit,
  `r vlink("a3-reading-and-exporting-results", "A3")` covers the accessors and
  export, and `r vlink("a4-figures", "A4")` builds the four signature figures.
- **Calibration in full** —
  `r vlink("m5-two-level-and-calibration", "M5")` gives the complete calibration
  scope (and the two-level surface that `GROUPS_ERROR` points to), including how
  `gp_calibrate()` relates to the deferred Table E1 Monte-Carlo.
- **The grading the solver produces** —
  `r vlink("m4-grading-frontier-and-report-cards", "M4")` derives the grade
  integer program, the frontier, and the report card;
  `r vlink("m1-foundations-and-notation", "M1")` fixes the notation.

The honest bottom line, restated: Gurobi is the recommended default and HiGHS the
open fallback; a non-optimal solve is never described as optimal; `gp_calibrate()`
is a diagnostic, not an E1 reproduction; and the one-level race and gender core is
accepted within its declared scope.

### Provenance

```{r provenance}
sessionInfo()
```

## References
