---
title: "M1: Foundations and notation"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This is the first method-track vignette: it pins down the notation every later
  vignette relies on, before any stage is derived. We state the KRW decision
  problem — grade noisy firm-level estimates by their latent effect, honestly,
  respecting the noise — and lay out a compact symbol table. We then fix the two
  load-bearing frozen invariants. The first is the grade-direction triple: why
  the exported grade 1 is the most-extreme theta, why the code carries the
  win-count S_i built from the binaries D_ij (not D_ji), and why the export is
  dense_rank(S_i, ascending = FALSE). The second is the firms-vs-names welfare
  flip: the identical machinery means opposite welfare readings on the same
  printed label, which is exactly why gradepath prints are result-first and
  quantity-anchored (the package decision GP-DEC-14-A). Each statement is verified in
  code against the bundled 97-firm
  parity fit, so the notation is not asserted but checked. The reader leaves with
  every symbol fixed and ready for m2-m5.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M1: Foundations and notation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

This is the foundations vignette for the method track. It is a lookup, not a
derivation: it summarizes the theory from Kline, Rose & Walters
[@kline2024discrimination] (the formal statements are in their Appendix A) and
pins the notation that m2-m5 build on. The two frozen invariants it fixes — the
grade-direction triple and the firms-vs-names welfare flip — are reproduced
faithfully and then checked in code, every displayed equation followed by a
verification panel against the bundled parity fit.

We load that fit once, at the top, and every panel below is a fast read or
compute off it. There is no solve.

```{r load-fit}
fit <- gp_parity_fit("race")
```

This is the proven 97-firm RACE report card: grades `2 / 81 / 14` at the KRW
baseline `lambda = 0.25`. The applied track shows how it is produced; here it is
the ground truth we check the notation against.

## 1. The KRW decision problem

You hold `n` units — here 97 large U.S. firms — each with a noisy point estimate
$\hat\theta_i$ of a latent effect $\theta_i$ and a standard error $s_i$ that says
how precisely $\hat\theta_i$ was measured. For firms, $\theta_i$ is the firm's
discrimination: the gap in callback rates between otherwise-identical White and
Black applicants. The estimates are heteroskedastic — some firms were measured
with far more applications than others — so the noisiest estimates look the most
extreme purely by chance.

> "Estimates of firm-level discrimination are inherently noisy ... A report card
> summarizes this evidence by sorting firms into a small number of grades."
> — Kline, Rose & Walters [@kline2024discrimination]

The decision problem is therefore not "rank the units" but "sort the units into a
few **honest grades** that respect the noise." Two units belong in the same grade
when the data cannot decisively separate them; they belong in different grades
only when the posterior evidence supports an ordering. The machinery that does
this — deconvolve a prior, shrink to posteriors, compare every pair, solve a
grade integer program along an information-reliability frontier — is empirical
Bayes in lineage [@walters2024empirical] and is the subject of m2-m5. This
vignette fixes the symbols those stages speak in, and the two directional facts
that are easiest to get backwards.

## 2. The symbol table

The table below is the compact reference: each symbol, the scale it lives on,
and a sign-or-direction note. It is reproduced from the package notation source
of truth (KRW Appendix A); downstream vignettes cite it rather than redefining.
The symbols group by the stage they first appear in — estimation,
standardization, prior/posterior, social choice, grading.

| Symbol | Meaning | Scale / units | Sign or direction note |
|---|---|---|---|
| $\theta_i$ | Latent unit effect: a firm's discrimination, or a name's contact propensity | Latent (standardized) | Larger $\theta$ = more extreme. *What "extreme" means flips by dataset — see §6.* |
| $\hat\theta_i$ | The noisy stage-1 estimate of $\theta_i$ | Same scale as $\theta$ | Input to standardization; not graded directly. |
| $s_i$ | Standard error / noise scale of $\hat\theta_i$ | Same scale as $\theta$ | Heteroskedastic across units; drives the precision dependence in §3. |
| $\beta$ | Precision-dependence exponent in $\theta = s^{\beta}\,v$ | Dimensionless | gradepath $\beta$-GMM estimate; $\beta = 0$ recovers the homoskedastic case. |
| $v_i$ | Precision-independent component, $v_i = \theta_i / s_i^{\beta}$ | Standardized | The latent quantity the prior $G$ models. |
| $r_i$ | Standardized residual scale, $r_i = \hat\theta_i / s_i^{\beta}$ | Standardized | **The scale on which the prior $G$ and the weights live.** Built in §3. |
| $G$ | The deconvolved prior over $v$ (equivalently over $r$) | Distribution | Estimated by gradepath's native deconvolution (m3). |
| $\pi_{ij}$ | Pairwise outranking posterior, $\pi_{ij} = \Pr(\theta_i > \theta_j \mid \text{data})$ | Probability $\in [0,1]$ | $\pi_{ij} + \pi_{ji} = 1$ (ties measure-zero). The primitive the grade IP consumes. §4. |
| $S_i$ | Code win-count, $S_i = \sum_j D_{ij}$ | Integer $\ge 0$ | The row-sum the **exported** grade is built from. §5. |
| $\text{grade}_i$ | The exported integer grade label, via $\operatorname{dense\_rank}(S_i,\ \text{asc}=\text{FALSE})$ | Integer $\in \{1,\dots,n\}$ | **Grade 1 = largest $S_i$ = most-extreme $\theta$.** Carries no superiority statement. §5. |
| $\mathrm{DR}$ | Discordance (disagreement) rate of a grading | Rate $\in [0,1]$ | Fraction of comparable pairs the grading orders against the posterior. Lower is cleaner. |
| $\bar\tau$ | Posterior expected Kendall rank agreement | $\in [-1,1]$ | Higher is cleaner; the information axis of the frontier. |
| $\lambda$ | Grade-coarseness / loss-tradeoff weight | $\in [0,1]$ | Separation threshold $1/(1+\lambda)$; baseline $\lambda = 0.25 \Rightarrow 0.80$. §7. |

Table: Every symbol the method track uses, defined once.

These symbols map onto concrete accessor slots on a fit, which is what lets us
verify each definition rather than assert it. The crosswalk we lean on below:
$\beta$ is `fit$precision_fit$parameters$beta`; the pairwise matrix
$\pi_{ij}$ is `get_pairwise(fit)$matrix`; the exported grade is
`as.data.frame(fit)$grade` (equivalently `get_grades(fit)`); the frontier
quantities $\mathrm{DR}$ and $\bar\tau$ live in `fit$grade_path$summary`; and the
selected $\lambda$ is `fit$selected_grade$lambda`.

## 3. Precision dependence and the r-scale

The first stage standardizes away the precision dependence. KRW's convention is

$$
\theta_i \;=\; s_i^{\beta}\, v_i,
$$

where $v_i$ is the precision-independent component and the exponent $\beta$ is
estimated by gradepath's $\beta$-GMM step. The object the rest of the pipeline
actually lives on is the **standardized r-scale**,

$$
r_i \;=\; \hat\theta_i \,/\, s_i^{\beta},
$$

the scale on which the prior $G$ and the comparison weights are built. The
derivation — the GMM moments, the multiplicative-vs-additive split — is m2; here
we only fix $\beta$ and $r$ and check them.

**Verification panel.** Read $\beta$ off the fit (it should match the published
KRW report), then compute the r-scale by hand from $\hat\theta$ and $s$. Note the
standard-error column is `est$s` (not `est$se`):

```{r verify-beta-r}
beta <- fit$precision_fit$parameters$beta
round(beta, 4)                              # the precision exponent

est <- fit$estimates
r   <- est$theta_hat / est$s^beta           # the standardized r-scale
head(r, 3)
```

gradepath lands on $\beta = `r round(beta, 4)`$, matching the `0.510` KRW report.
The same value is reachable through the prior accessor, `get_prior(fit)$metadata$beta`,
since the prior is the object that lives on the r-scale. The first three r-scale
values are `r paste(round(head(r, 3), 4), collapse = ", ")` — there is no
`fit$precision_fit$r` slot; the r-scale is computed, never stored, exactly as
above.

## 4. The pairwise primitive

Before grades there are **pairwise outranking posteriors**. For every ordered
pair $(i, j)$,

$$
\pi_{ij} \;=\; \Pr\!\big(\theta_i > \theta_j \mid \text{data}\big),
\qquad
\pi_{ij} + \pi_{ji} = 1,
$$

with ties measure-zero under the continuous prior $G$. This matrix — the
posterior probability that one unit's latent effect exceeds another's — is the
only object the grade integer program consumes. m3 derives the posteriors and m4
shows how the IP reads $\pi$.

**Verification panel.** Pull the pairwise matrix and confirm the antisymmetry
identity holds exactly: a single entry plus its transpose is 1, and the maximum
deviation across all off-diagonal pairs is 0.

```{r verify-pi-antisymmetry}
M <- get_pairwise(fit)$matrix
M[1, 2] + M[2, 1]                           # one pair: exactly 1

max(abs((M + t(M))[upper.tri(M)] - 1))      # worst-case deviation across all pairs
```

The pair sums to `r M[1, 2] + M[2, 1]` and the worst-case deviation is
`r max(abs((M + t(M))[upper.tri(M)] - 1))`, so $\pi_{ij} + \pi_{ji} = 1$ holds to
machine zero. This is the primitive the grade label is built from next.

## 5. The grade-direction triple

This is the most parity-critical definition in the package, and the one most
easily gotten backwards. Three different objects each carry a notion of "rank,"
they use **different index conventions**, and they point in **different
directions**. Stating all three together, once, is the only defense against an
off-by-transpose error that would silently invert every report card.

**1. Paper ordinal (top-is-1).** KRW assign

$$
d_i^{*} \;=\; 1 + \sum_{j} d_{ji}^{*},
$$

where $d_{ji}^{*} = 1$ when unit $j$ sits above unit $i$ in the partial order. So
**paper grade 1 is the TOP** of the order (nothing above it), and the grade
*number* increases as a unit falls. Note the transposed index order $d_{ji}^{*}$:
it counts how many units are above $i$.

**2. Code win-count (independent binaries).** The solver works with

$$
S_i \;=\; \sum_{j} D_{ij},
$$

where $D_{ij} = 1$ when the grading ranks $i$ **above** $j$. The code uses
$D_{ij}$, **not** $D_{ji}$, and $D_{ij}$ and $D_{ji}$ are **independent** solver
binaries (the IP does not hard-impose $D_{ij} + D_{ji} = 1$; antisymmetry and
transitivity enter as constraints, see m4). So $S_i$ is a *win-count*: how many
units $i$ is placed above.

**3. Exported grade (dense-rank of the win-count, descending).** The grade
gradepath exports is

$$
\text{grade}_i \;=\; \operatorname{dense\_rank}\!\big(S_i;\ \text{ascending}=\text{FALSE}\big),
$$

so **exported grade 1 = the LARGEST $S_i$ = the most-extreme $\theta$.** The
largest win-count earns the label 1.

The three reconcile by design. The unit at the top of the partial order has
nothing above it, so it is placed above the most other units and earns the
largest $S_i$ — and therefore the export's label 1. The paper's top-is-1 ordinal
and the code's largest-$S_i$-is-1 export thus agree on *which* unit gets the
label 1: the most-extreme one.

> **The one sentence to remember.** Exported grade 1 = largest $S_i$ = the
> most-extreme $\theta$; the code uses $D_{ij}$ (not $D_{ji}$); the export is
> `dense_rank(S_i, ascending = FALSE)`. The grade label is an integer in
> $\{1, \dots, n\}$ — a sorting index, not a letter such as A or B — and carries
> no superiority statement.

**Verification panel.** We cannot read the internal $D_{ij}$ or $S_i$ off the
fit (they are solver-internal), but the directional consequence is checkable on
the exported label: if grade 1 is the most-extreme $\theta$, then the mean
posterior effect must be **largest** in grade 1 and decrease monotonically as the
grade number rises. Group the report card's `posterior_mean` by grade:

```{r verify-grade-direction}
card <- as.data.frame(fit)
tapply(card$posterior_mean, card$grade, mean)
```

Grade 1 has the highest mean posterior effect (`r round(tapply(card$posterior_mean, card$grade, mean)[["1"]], 3)`),
grade 2 is lower (`r round(tapply(card$posterior_mean, card$grade, mean)[["2"]], 3)`),
and grade 3 is lowest (`r round(tapply(card$posterior_mean, card$grade, mean)[["3"]], 3)`)
— monotonically decreasing. This numerically confirms that the exported grade 1
is the most-extreme $\theta$, exactly as the triple states.

## 6. The firms-vs-names welfare flip

The grading machinery is *identical* for both KRW datasets. What differs is the
**welfare meaning** of "most extreme," and therefore of the printed grade 1. This
is the second frozen invariant, and the reason result prose must be written
carefully. On the **same exported label** produced by the dense-rank rule of §5:

- **FIRMS.** Grade 1 = the **most discriminatory** firm (largest $\theta$ = the
  widest gap in callback by applicant race). This is the worst firm for workers.
  Directional prose: *"more discriminatory,"* *"a wider contact-rate gap."*
- **NAMES.** Grade 1 = the **best-treated** name (largest $\theta$ = the highest
  contact/callback rate). This is the best outcome for that name. Directional
  prose: *"a higher posterior contact rate."*

The machinery does **not** know which dataset it is grading; the flip is in the
*interpretation*, supplied by the analyst, not in the code. So one must **never**
write a valenced verb that inverts in meaning between firms and names — generic
"best/worst" across both datasets inverts, as do generic ranking-contest verbs.
Only the directional, quantity-anchored prose above is safe: it names the
quantity ("a wider contact-rate gap," "a higher posterior contact rate") rather
than a verdict, so the same sentence reads correctly for either dataset.

This is exactly why gradepath's print methods and report-card labels follow the
package decision `GP-DEC-14-A`: the rendered output is **result-first** and names
the quantity (a posterior gap, a contact rate) rather than a valenced verdict, so
the same code path serves both datasets without the printed text lying about one
of them.
The package ships the firms data; names is a separate KRW dataset, so the flip
here is conceptual — but the discipline it imposes on the prose is in force on
every fit. No numeric panel is needed: the grade label is computed identically
for both datasets, and the welfare reading is the analyst's to supply.

## 7. Loss, risk, and the lambda threshold

A light note, with the full treatment deferred to m4. The grade integer program
minimizes a Bayes risk that trades the discordance rate $\mathrm{DR}$ against the
posterior Kendall agreement $\bar\tau$, governed by the coarseness weight
$\lambda \in [0, 1]$. The one piece of $\lambda$ mechanics worth fixing here is
the **separation threshold**: a pair $(i, j)$ may share a grade only when neither
outranks the other decisively, i.e. when

$$
\max(\pi_{ij},\, \pi_{ji}) \;<\; \frac{1}{1+\lambda}.
$$

At the baseline $\lambda = 0.25$ the threshold is $1/1.25 = 0.80$: two units may
share a grade only if the stronger pairwise probability is below $0.80$. At
$\lambda = 1$ the threshold is $0.5$ and the partition collapses to the Condorcet
(finest-separation) ordering. The bundled fit was selected at the baseline, which
we can read directly:

```{r verify-lambda}
fit$selected_grade$lambda                   # the selected coarseness weight
1 / (1 + fit$selected_grade$lambda)         # the separation threshold
```

The fit sits at $\lambda = `r fit$selected_grade$lambda`$ with separation
threshold $`r round(1 / (1 + fit$selected_grade$lambda), 2)`$. The grade IP that
turns this threshold into the partition — and the frontier the package sweeps
$\lambda$ along — is derived in m4.

## 8. Where to next

With the symbols fixed and the two invariants checked, the stage derivations
follow in order:

- **Precision and standardization** — `r vlink("m2-precision-and-standardization", "m2")`
  derives the $\beta$-GMM step behind §3 (the moments, the multiplicative-vs-additive split).
- **Deconvolution and posterior** — `vignette("m3-deconvolution-and-posterior")`
  builds the prior $G$, the posteriors, and the pairwise matrix $\pi_{ij}$ of §4.
- **Grading frontier and report cards** — `vignette("m4-grading-frontier-and-report-cards")`
  derives the grade IP, the win-count $S_i$ of §5, and the $\lambda$ frontier of §7.
- **Two-level and calibration** — `vignette("m5-two-level-and-calibration")`
  extends the model to industries and validates the pipeline by Monte Carlo.

For the workflow side, the applied track runs the same pipeline as a hands-on
example: `r vlink("a1-getting-started", "a1")` is the first contact, and
`vignette("a2-the-grading-workflow")` walks the seven stages with race and gender
side by side.

### Provenance

```{r provenance}
sessionInfo()
```

## References
