---
title: "M2: Precision and standardization"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This is the beta-GMM precision core of the KRW pipeline, made rigorous.
  Heteroskedastic stage-1 estimates make the noisiest units look the most
  extreme; KRW standardize this away with the precision-dependence model
  theta = s^beta v, estimating the precision exponent beta by two-step
  optimally-weighted GMM rather than assuming it. This vignette derives the
  moment conditions and the GMM, explains the multiplicative (race) versus
  additive (gender) split, and verifies beta = 0.5095 (race) and 1.2554
  (gender) live against the bundled parity fits, with the GMM overidentification
  J-test. The reader leaves with the standardized r-scale the deconvolution of
  m3 operates on.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M2: Precision and standardization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

This is the second method-track vignette, and the first to derive a stage. It
takes the one paragraph of `r vlink("m1-foundations-and-notation", "m1")`'s §3 —
the line that said "the precision dependence is standardized away, the
derivation is m2" — and makes it rigorous. We summarize the theory from Kline,
Rose & Walters [@kline2024discrimination] (their §5.1, Equations 8–10) rather
than re-derive every algebra step, and we follow every displayed equation with a
verification panel that confirms it live against **both** bundled parity fits.

We load both fits once, at the top. Every panel below is a fast read off an
already-fitted `precision_fit` slot — there is no GMM solve anywhere in this
document. The fits are pre-solved, so the beta-GMM stage we are deriving has
already run; here it is the ground truth we check the precision core against.

```{r load-fits}
fit_race   <- gp_parity_fit("race")     # 97-firm RACE card, grades 2 / 81 / 14
fit_gender <- gp_parity_fit("gender")   # 97-name GENDER card, grades 1 / 3 / 89 / 4
```

These are the 97-firm RACE and gender parity fits at the KRW baseline
`lambda = 0.25`. The applied track (`r vlink("a2-the-grading-workflow", "a2")`)
shows how they are produced. The precision fit lives in one slot,
`fit$precision_fit`, with the same layout on both demographics — the crosswalk we
verify against throughout:

```{r show-precision-slots}
names(fit_race$precision_fit)
```

`parameters` carries the estimated model (`model_form`, `beta`, `mu`,
`characteristic`); `moments` carries the GMM moment vector `m_hat`, its
covariance `V_m`, and the overidentification test (`J`, `df`, `p_value`); and
`scale` records that the prior lives on the standardized r-scale.

## 1. Why standardize

Start with the problem, because the whole stage exists to solve it. You hold $n$
units — 97 large U.S. firms — each with a noisy stage-1 estimate $\hat\theta_i$
of a latent effect $\theta_i$ and a standard error $s_i$ that says how precisely
$\hat\theta_i$ was measured. The estimates are **heteroskedastic**: some firms
were measured with far more correspondence applications than others, so their
$s_i$ varies by an order of magnitude across the panel.

Heteroskedasticity has a sharp consequence for any procedure that compares units
by how extreme they look. A firm measured with few applications has a large
$s_i$, and a large $s_i$ means $\hat\theta_i$ can land far from $\theta_i$ by
chance alone. So the noisiest units are *exactly* the ones most likely to
produce an eye-catching $\hat\theta_i$ — not because their latent effect is
extreme, but because their estimate is loose. A grading that took $\hat\theta_i$
at face value would hand the most-extreme grades to the least-precisely-measured
firms, which is precisely backwards.

> "Estimates of firm-level discrimination are inherently noisy ... A report card
> summarizes this evidence by sorting firms into a small number of grades."
> — Kline, Rose & Walters [@kline2024discrimination]

There is a second, subtler reason, and it is the one KRW build the model around.
Empirical Bayes shrinkage [@walters2024empirical] borrows strength across units
by shrinking each estimate toward a common prior, with the amount of shrinkage
set by each unit's noise. That machinery assumes the latent effect $\theta_i$ is
drawn *independently* of its standard error $s_i$. In the firm setting that
assumption is implausible: firms with many postings generate more applications
(small $s_i$) and may also differ systematically in their discrimination. If
$\theta_i$ and $s_i$ are correlated, a single prior fit to the whole sample
over-shrinks some units and under-shrinks others, and the bias is invisible — it
hides inside the very precision the method relies on.

KRW's fix is to put every unit on a **common precision scale** before any
comparison or shrinkage happens. The deconvolution of
`r vlink("m3-deconvolution-and-posterior", "m3")`, the pairwise posteriors, and
the grading frontier of
`r vlink("m4-grading-frontier-and-report-cards", "m4")` all operate on that
scale — never on the raw $\hat\theta_i$. The rest of this vignette derives the
scale and checks the one number that defines it.

## 2. The precision-dependence model and the r-scale

The model is a power law linking the latent effect to the precision of its
measurement [@kline2024discrimination, Eq. 8]:

$$
\theta_i \;=\; s_i^{\beta}\, v_i,
\qquad
v_i \mid s_i \;\overset{\text{iid}}{\sim}\; G_v .
$$

Read it as a decomposition. The factor $s_i^{\beta}$ carries everything about
$\theta_i$ that travels with precision; the residual $v_i = \theta_i/s_i^{\beta}$
is what is left once that dependence is divided out, and *by construction* $v_i$
is conditionally independent of $s_i$. The exponent $\beta$ is the one number
that controls how much the latent effect rides on precision:

- $\beta = 0$ gives $\theta_i = v_i$ — the homoskedastic textbook case, no
  precision dependence, latent effect independent of standard error.
- $\beta \approx 0.5$ is the square-root-of-precision regime: the latent scale
  grows like the standard deviation of the noise, halfway between "no
  dependence" and "fully proportional."
- $\beta = 1$ makes the latent scale move one-for-one with $s_i$.

The crucial methodological point — the reason this stage is GMM and not a guess —
is that $\beta$ is **estimated, not assumed**. Picking $\beta = 0$ (assume away
the problem) or $\beta = 1$ (assume it fully) would each bias every downstream
residual, prior, posterior, and grade. The data are allowed to say where between
those poles the truth sits.

Once $\beta$ is in hand, the object the rest of the pipeline lives on is the
**standardized r-scale**: the residual you get by dividing the *estimate* by its
own $s_i^{\beta}$,

$$
r_i \;=\; \hat\theta_i \,/\, s_i^{\beta} .
$$

This is the scale on which the prior $G$ is deconvolved (m3) and the comparison
weights are built. It is also where the noise rescales: under the model the
estimate is $\hat\theta_i = \theta_i + \varepsilon_i$ with
$\varepsilon_i \sim N(0, s_i^2)$, so dividing by $s_i^{\beta}$ leaves the residual
with noise scale $s_{v,i} = s_i^{1-\beta}$. That $s_{v,i} = s_i^{1-\beta}$ is the
*likelihood* scale m3 will use; keep it in view, because the moments in §3 are
built from it.

**Verification panel.** Read $\beta$ off the race fit, then build the r-scale by
hand from $\hat\theta$ and $s$. Note the standard-error column is `est$s` (not
`est$se`), and there is no stored `precision_fit$r` slot — the r-scale is
*computed*, never stored:

```{r verify-beta-r}
beta_race <- fit_race$precision_fit$parameters$beta
round(beta_race, 4)                           # the race precision exponent

est <- fit_race$estimates
r   <- est$theta_hat / est$s^beta_race        # the standardized r-scale
head(r, 3)
```

gradepath lands on $\beta = `r round(beta_race, 4)`$ for race, a faithful match
to the $\sim 0.510$ KRW report. The same value is reachable through the prior
accessor, `get_prior(fit_race)$metadata$beta`, because the prior is the object
that lives on the r-scale. The first three r-scale values are
`r paste(round(head(r, 3), 4), collapse = ", ")`.

## 3. The moment conditions

Why is $\beta$ identifiable at all? Because the model makes a *testable* promise:
after standardizing, the residual should look the same regardless of how
precisely it was measured. KRW turn that promise into moment conditions
[@kline2024discrimination, Eq. 9].

Form the **studentized residual** — the r-scale point estimate recentered and
scaled to be mean-zero, unit-variance under the model. With
$\hat v_i = \hat\theta_i/s_i^{\beta}$, its noise scale $s_{v,i} = s_i^{1-\beta}$,
and the implied latent SD $\sigma$,

$$
r_i \;=\; \frac{\hat v_i - \mu}{\sqrt{\sigma^2 + s_{v,i}^2}}
\quad\text{(race)},
\qquad
r_i \;=\; \frac{\hat v_i}{\sqrt{\sigma^2 + s_{v,i}^2}}
\quad\text{(gender, with }\hat v_i = (\hat\theta_i - \mu)/s_i^{\beta}\text{)} ,
$$

so that under correct specification $r_i$ is approximately i.i.d. with mean zero
and unit variance — *regardless of $s_i$*. (The two forms — subtracting $\mu$ in
the numerator versus inside $\hat v_i$ — are the multiplicative-versus-additive
split of §6; the four moments below are identical for both.) The four base moment
conditions ask exactly the mean-zero, unit-variance, $s$-free property, in two
pairs:

$$
\underbrace{E[r_i] = 0}_{\text{mean is zero}},
\quad
\underbrace{E[r_i\, s_i] = 0}_{\text{...even conditional on }s_i},
\quad
\underbrace{E[r_i^2 - 1] = 0}_{\text{variance is one}},
\quad
\underbrace{E[(r_i^2 - 1)\, s_i] = 0}_{\text{...even conditional on }s_i},
$$

stacked into the sample vector
$g_i = [\,r_i,\; r_i s_i,\; r_i^2 - 1,\; (r_i^2 - 1) s_i\,]$.

The structure *is* the identification story, and it is worth slowing down on. The
two **level** moments ($r$ and $r^2-1$) pin the location $\mu$ and scale $\sigma$
of the standardized residual. The two **interactions with $s_i$** ($r\,s$ and
$(r^2-1)\,s$) are the ones that earn their keep: they test whether the residual's
mean or variance still trends with precision *after* standardizing. If $\beta$
were wrong, some precision dependence would survive into the residual, those
interaction moments would not vanish, and the GMM would move $\beta$ to make them
vanish. That residual-versus-$s$ dependence is precisely what identifies $\beta$
*separately* from $\sigma$ — which is why this is GMM and not a conditional-mean
least-squares fit. A mean regression has no analogue of the variance moments and
so cannot recover $\beta$ this way.

These four moments identify the one-level parameter vector
$\delta = [\mu,\ \log\sigma_\xi,\ \beta]$ — **three parameters, four moments**, so
**one over-identifying restriction** is left over and becomes the specification
test of §5. (The intercept enters differently under the additive gender form, but
the count still leaves the same single over-identifying degree of freedom — which
is what the J-test reads for both fits in §5.) The moment **covariance** is
estimated cluster-robust on industry — one-level it is the ordinary
outer-product average $\hat\Omega = \tfrac1N\sum_i g_i g_i'$; with industry
structure it sums cross-products only within shared industries — so the standard
errors respect the within-industry coupling. The two-level extension that appends
two between-industry moments is `r vlink("m5-two-level-and-calibration", "m5")`;
here we use the four base moments. This stored moment vector and its covariance
are not an abstraction — they are slots on the fit, which is what lets us verify
the GMM rather than assert it.

## 4. Two-step optimally-weighted GMM

GMM chooses $\delta = [\mu,\ \log\sigma_\xi,\ \beta]$ to make the sample moment
average $g(\delta)$ as close to zero as a weight matrix $\mathsf W$ allows,
minimizing the quadratic form

$$
J(\delta;\,\mathsf W) \;=\; N\, g(\delta)'\,\mathsf W\, g(\delta) .
$$

The choice of $\mathsf W$ is the whole game with an over-identified system,
because with four moments and three parameters you cannot drive all four to zero
at once — you must decide which to favor. The **efficient** answer, standard GMM
theory, is to weight each moment by the inverse of its variance: set
$\mathsf W = \hat\Omega^{-1}$, so noisy moments count for less. But $\hat\Omega$
itself depends on $\delta$, which is the chicken-and-egg the **two-step**
procedure resolves:

1. **First step ($\mathsf W = I$).** Minimize $J(\delta; I)$ with the identity
   weight to get a consistent first-pass $\hat\delta^{(1)}$. KRW seed the search
   at their published estimates (race one-level $\approx [-1.18,\ -1.58,\ 0.51]$)
   and otherwise multistart.
2. **Reweight.** Evaluate $\hat\Omega_1 = \hat\Omega(\hat\delta^{(1)})$ and set the
   optimal weight to its inverse, $\mathsf W = \hat\Omega_1^{-1}$ (implemented as
   the Cholesky factor of $\hat\Omega_1^{-1}$, so the objective is the squared
   norm $\lVert \mathsf W^{1/2} g\rVert^2$ the optimizer minimizes directly).
3. **Second step.** Re-minimize $J(\delta; \mathsf W)$ from $\hat\delta^{(1)}$ to
   get the reported $\hat\delta$. Standard errors come from the optimal-GMM
   sandwich $C = (G'\hat\Omega_2^{-1} G)^{-1}/N$ with
   $G = \partial g/\partial\delta'$; because $\hat\Omega$ was clustered on
   industry, the SEs are cluster-robust.

The intuition for "optimal weight = inverse covariance" is the same one behind
weighted least squares: a moment you can measure tightly deserves a tight
constraint, and a moment drowning in sampling noise should not be allowed to drag
the estimate around. Inverse-variance weighting is just that instinct made
precise — and it is also what makes the leftover quadratic at the optimum into a
clean $\chi^2$ statistic (§5).

From this *single* fit the core emits two carriers the deconvolution of m3
consumes: the matched **moment vector** $\hat m$ (the GMM-implied prior location
and scale) and its delta-method **covariance** $V_m$. These are what m3 scores
its penalty grid and builds its support caps against, which is why the precision
and deconvolution steps are one unified core rather than two independent fits.

**Verification panel.** Read $\hat m$ and $V_m$ off the race fit — what GMM
minimized is encoded right here:

```{r verify-moments}
mom <- fit_race$precision_fit$moments
mom$m_hat                                      # GMM-implied [mu, sigma_xi]
round(mom$V_m, 6)                              # their (delta-method) covariance
isSymmetric(mom$V_m)                           # symmetric: TRUE
eigen(mom$V_m, symmetric = TRUE, only.values = TRUE)$values  # both > 0 => positive-definite
```

The implied prior mean is
$\hat m_1 = `r round(fit_race$precision_fit$moments$m_hat[1], 4)`$ — exactly the
$\mu$ in `parameters` (the race intercept) — and the implied scale is
$\hat m_2 = `r round(fit_race$precision_fit$moments$m_hat[2], 4)`$. The covariance
$V_m$ is symmetric (so `isSymmetric()` returns `TRUE`) with both eigenvalues
positive — symmetric positive-definite, exactly as a delta-method covariance must
be.
Together they are the coupling the next stage's penalty grid and support caps are
built from; here they confirm the GMM emits both from a single fit.

## 5. The overidentification J-test

With four moments and three parameters, one restriction is over-identifying — and
that surplus is a free specification test, no extra data required. At the
optimum, the weighted quadratic does not quite reach zero (it cannot, with a
surplus moment), and **how far it falls short** is the test statistic. Hansen's
J-statistic is the second-step objective at the optimum,

$$
J \;=\; N\, g(\hat\delta)'\,\hat\Omega^{-1} g(\hat\delta)
\;\overset{a}{\sim}\; \chi^2(\#\text{moments} - \#\text{params}),
$$

with one degree of freedom here ($4 - 3 = 1$; the $\chi^2(1)$ critical value at
the 5% level is $3.84$). Read it plainly: if the precision model is right, the
standardized residual really is mean-zero, unit-variance, and free of
$s$-dependence, so all four moments are near zero, $J$ is small, and the p-value
is large. A large $J$ (small p-value) would say the moments *cannot* be jointly
satisfied — the $\theta = s^{\beta} v$ form is too rigid for the data. A $J$ that
grew with the sample size would be the signature of a misspecified model, not
sampling noise.

**Verification panel.** Read the J-statistic, its degrees of freedom, and its
p-value off both fits:

```{r verify-jtest}
jt <- function(fit) {
  m <- fit$precision_fit$moments
  c(J = round(m$J, 4), df = m$df, p = round(m$p_value, 4))
}
rbind(race = jt(fit_race), gender = jt(fit_gender))
```

For race, $J = `r round(fit_race$precision_fit$moments$J, 4)`$ on
`r fit_race$precision_fit$moments$df` degree of freedom, p-value
`r round(fit_race$precision_fit$moments$p_value, 4)`; for gender,
$J = `r round(fit_gender$precision_fit$moments$J, 4)`$, p-value
`r round(fit_gender$precision_fit$moments$p_value, 4)`. Both p-values are far
above any conventional threshold, so **the moment conditions are not rejected**:
the precision-dependence model is consistent with the data for both
demographics. That is the honest reading — the J-test does not establish the
model, it fails to find evidence against it, which is exactly the assurance you
want before deconvolving on the r-scale.

## 6. The multiplicative-versus-additive split

The model in §2 was written multiplicatively: $\theta_i = s_i^{\beta} v_i$, with
the residual formed as $\hat v_i = \hat\theta_i / s_i^{\beta}$. That is the race
form. Gender uses an **additive** variant that subtracts a location *before*
dividing:

$$
\text{multiplicative (race):}\quad
\hat v_i = \frac{\hat\theta_i}{s_i^{\beta}},
\qquad
\text{additive (gender):}\quad
\hat v_i = \frac{\hat\theta_i - \mu}{s_i^{\beta}} .
$$

Why two forms? Because the location of the latent effect means different things
for the two demographics, and the model's intercept $\mu$ has to sit where the
data centers. For **race**, the average firm discriminates against minority
applicants, so the latent location is bounded away from zero and *positive* — KRW
parameterize it multiplicatively (internally $\mu = e^{\delta_1}$, which keeps it
positive), and the standardization scales around that positive center. For
**gender**, the average gap can fall on either side of zero, so the location is
modeled as a free additive intercept $\mu$ that is subtracted off before
standardizing. The split is not a tuning knob the analyst turns: gradepath
**selects the form per demographic** to match the centering KRW use, and the
choice propagates through the moment conditions and the variance formulas.

**Verification panel.** Read the model form, the precision exponent, and the
intercept for both fits side by side:

```{r verify-split}
split_row <- function(fit) {
  p <- fit$precision_fit$parameters
  c(model_form = p$model_form,
    beta = round(p$beta, 4),
    mu   = round(p$mu, 4))
}
rbind(race = split_row(fit_race), gender = split_row(fit_gender))
```

Race comes back **multiplicative** with
$\beta = `r round(fit_race$precision_fit$parameters$beta, 4)`$ and a positive
intercept $\mu = `r round(fit_race$precision_fit$parameters$mu, 4)`$; gender comes
back **additive** with
$\beta = `r round(fit_gender$precision_fit$parameters$beta, 4)`$ and a near-zero
intercept $\mu = `r round(fit_gender$precision_fit$parameters$mu, 4)`$. The
positive race intercept and the near-zero gender intercept are exactly the
centering story above, read straight off the fits. Note too that the two $\beta$
values differ substantially ($\approx 0.51$ vs $\approx 1.26$): precision
dependence is genuinely stronger in the gender panel, which is another reason
$\beta$ must be estimated separately rather than fixed at a shared value.

## 7. What the next stage receives

The precision core hands one scale and two carriers forward. The **scale** is the
r-scale: the prior $G$ and the comparison weights are deconvolved and built on
$r_i = \hat\theta_i / s_i^{\beta}$, with the per-unit likelihood noise
$s_{v,i} = s_i^{1-\beta}$. The **carriers** are the matched moment vector $\hat m$
and its covariance $V_m$ from §4, which m3's deconvolution uses to score its
penalty grid and set its support caps. Everything downstream — the deconvolved
prior, the posteriors, the pairwise outranking matrix, the grades — lives on this
r-scale and never sees the raw $\hat\theta_i$ again.

**Verification panel.** Confirm the recorded scale, and rebuild the gender
r-scale in its additive form — `(theta_hat - mu)/s^beta`, using the gender fit's
own $\beta$ and intercept $\mu$ (the additive variant of §6, not the
multiplicative race transform):

```{r verify-scale}
fit_race$precision_fit$scale                  # the scale the prior lives on: "r"

est_g  <- fit_gender$estimates
beta_g <- fit_gender$precision_fit$parameters$beta
mu_g   <- fit_gender$precision_fit$parameters$mu
head((est_g$theta_hat - mu_g) / est_g$s^beta_g, 3)   # gender r-scale: additive form, (theta_hat - mu)/s^beta
```

The scale slot reads `r dQuote(fit_race$precision_fit$scale, q = FALSE)`, and the
gender r-scale is computed by the additive transform of §6 — subtracting the
gender intercept $\mu$ before dividing by $s^\beta$ — using the gender fit's own
$\beta$ and $\mu$. Because the gender intercept is near zero, this lands close to
the raw quotient, but the form is the additive one the deconvolution actually
receives. That single column is what
`r vlink("m3-deconvolution-and-posterior", "m3")` deconvolves a prior over.

## 8. Where to next

With $\beta$ estimated, the r-scale built, and the precision model checked
against both fits and its own J-test, the pipeline continues:

- **Foundations and notation** — `r vlink("m1-foundations-and-notation", "m1")`
  is the symbol table and the two frozen invariants this vignette assumed; its
  §3 is the one-paragraph version of everything above.
- **Deconvolution and posterior** — `vignette("m3-deconvolution-and-posterior")`
  takes the r-scale and the carriers $\hat m$, $V_m$ from here, deconvolves the
  prior $G$ over $v$, and forms the posteriors and the pairwise matrix
  $\pi_{ij}$.
- **Grading frontier and report cards** — `vignette("m4-grading-frontier-and-report-cards")`
  turns the posteriors into grades — integer labels in $\{1, \dots, n\}$, never
  letter grades — via the grade integer program and the $\lambda$ frontier.
- **Two-level and calibration** — `vignette("m5-two-level-and-calibration")`
  adds the two between-industry moments that extend §3 to the six-moment
  industry model, and validates the whole pipeline by Monte Carlo.

For the workflow side, `r vlink("a1-getting-started", "a1")` is the first
contact, `r vlink("a2-the-grading-workflow", "a2")` runs this same precision
stage hands-on with race and gender side by side, and
`r vlink("a5-solvers-and-calibration", "a5")` covers the calibration harness that
the J-test logic above feeds into. Throughout, gradepath's printed output is
result-first and quantity-anchored (the package decision GP-DEC-14-A): it names
the quantity rather than a verdict, so the same code path serves both
demographics without the printed text inverting on one of them.

### Provenance

```{r provenance}
sessionInfo()
```

## References
