---
title: "M3: Deconvolution and posterior"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This is the empirical-Bayes core of the KRW pipeline, stated rigorously. From
  the standardized r-scale that m2 hands forward, gradepath deconvolves a prior G
  over the latent effects — a log-spline density, with no parametric family
  assumed — that separates the true cross-unit spread from the measurement noise
  blurring each estimate. It then forms every unit's posterior by combining its
  noisy estimate with that prior, so the noisier a unit's estimate the harder its
  posterior shrinks toward the prior and the tighter its credible interval. From
  the posteriors it integrates the pairwise outranking matrix Pi, whose entry is
  the posterior probability that one unit's latent effect exceeds another's. This
  vignette states each step crisply and verifies it live on the bundled one-level
  parity fit: the prior integrates to one over its grid, the posterior means and
  standard deviations are demonstrably pulled in, and Pi is antisymmetric to
  machine zero with a one-half diagonal. The two-level industry extension — a
  tensor-product prior over a within-industry and a between-industry component,
  pushed forward to the theta-scale — is described here and verified in m5, since
  the fit shipped with the package is one-level.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M3: Deconvolution and posterior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

This is the third method-track vignette and the heart of the empirical-Bayes
machinery. `r vlink("m2-precision-and-standardization", "m2")` ended by handing
forward one scale — the standardized r-scale $r_i = \hat\theta_i / s_i^{\beta}$ —
and two carriers, the matched moment vector $\hat m$ and its covariance $V_m$.
This vignette takes that scale and derives the three objects the grading of
`r vlink("m4-grading-frontier-and-report-cards", "m4")` reads: the deconvolved
prior $G$, the empirical-Bayes posterior per unit, and the pairwise outranking
matrix $\pi_{ij}$. We summarize the theory from Kline, Rose & Walters
[@kline2024discrimination] and the empirical-Bayes lineage it sits in
[@robbins1956empirical; @efron2016empirical; @walters2024empirical] rather than
re-derive every line, and we follow each displayed equation with a verification
panel checked live against the bundled fit.

We load that fit once, at the top. Every panel below is a fast read off an
already-fitted slot — `get_prior`, `get_posterior`, or `get_pairwise` — so there
is **no deconvolution solve anywhere in this document**. The fit is pre-solved:
the deconvolution we are about to state has already run, and here it is the ground
truth we check the EB core against.

```{r load-fit}
fit <- gp_parity_fit("race")   # the proven 97-firm RACE card, grades 2 / 81 / 14
```

This is the 97-firm RACE parity fit at the KRW baseline `lambda = 0.25`. It is a
**one-level** fit — no industry structure — which is exactly the case this
vignette verifies end to end. The two-level industry extension is described in §5
and verified in `r vlink("m5-two-level-and-calibration", "m5")`.

## 1. From a standardized scale to a recovered world

The r-scale handed over by m2 is not yet what the grades are built on. Each
$r_i$ is a *noisy* read on the latent component $v_i$: standardizing by
$s_i^{\beta}$ removes the precision dependence but not the sampling blur. The
cross-sectional spread of the observed $r_i$ is therefore the spread of the true
latent effects **plus** the spread of the noise, combined in quadrature — wider
than the truth. Deconvolution is the step that separates the two.

Formally it asks: given the observed (blurred) distribution and the known per-unit
noise scale, what distribution of latent effects, once blurred, reproduces what we
see? The answer is the prior $G$ — the recovered distribution of the true effects
with the noise removed. This is the empirical-Bayes program in its classical form
[@robbins1956empirical; @efron2010large]: the ensemble of noisy estimates
identifies the prior they were drawn from even though no single estimate does.

Why the report card needs $G$ rather than the raw $r_i$ is the argument m2 made:
the noisiest units look the most extreme by chance, so grading the raw estimates
would hand the loudest grades to the least-certain units. $G$ is the corrective —
it lets each unit's posterior ask "given how noisy *this* estimate is, where does
the latent effect most likely sit?" The pipeline is therefore three steps, in
order: **deconvolve** the prior (§2), **shrink** to posteriors (§3), and
**compare** every pair (§4). The two-level industry extension layers a second
component onto $G$ and is the subject of §5.

## 2. The deconvolution problem

Write the standardized model from m2 as a measurement equation. On the r-scale
each observation is the latent component $v_i \sim G$ plus Gaussian noise of
per-unit scale $s_{v,i} = s_i^{\,1-\beta}$:

$$
r_i \;=\; v_i + \varepsilon_i,
\qquad
v_i \sim G,
\qquad
\varepsilon_i \,\mid\, s_i \;\sim\; N\!\big(0,\; s_{v,i}^2\big),
\qquad
s_{v,i} = s_i^{\,1-\beta} .
$$

The marginal distribution of the observed $r_i$ is the **convolution** of the
unknown prior $G$ with the per-unit normal noise; recovering $G$ from that marginal
is the deconvolution problem [@efron2016empirical; @walters2024empirical]. We do
not recover the latent draws $v_i$ one at a time — we recover the **distribution**
$G$ they came from, which is both better identified (the ensemble is informative
where a single estimate is not) and exactly what the posterior of §3 requires.

The defining choice is that $G$ is **not** assumed to belong to a parametric
family. gradepath follows KRW with a **penalized log-spline** prior: the density on
a fixed support grid is written as a softmax, $g_m \propto \exp(Q_m\,\alpha)$,
where $Q$ is a spline basis on the grid and $\alpha$ a short coefficient vector.
The softmax keeps the density positive and summing to one, and $\alpha$ is fit by
penalized maximum likelihood so the convolved $\hat g$ matches the observed spread.
The penalty is selected against the m2 carriers $\hat m$ and $V_m$ — which is why
the precision and deconvolution stages form one coupled core, not two independent
fits — and the support is capped from the GMM-implied scales rather than the data
maximum. gradepath's one-level deconvolution is **native**; its one-level slice is
cross-checked against `ebrecipe`'s `eb_deconvolve` as a textbook sanity gate, with
the KRW reference as the production parity target. The full derivation of the
penalty grid and the caps lives in
`r vlink("m5-two-level-and-calibration", "m5")`, where the two-level version is
built. The fact to carry here is that **no parametric family is assumed**: the
recovered $G$ can be skewed or heavy-tailed if the data demand it.

**Verification panel.** Pull the deconvolved prior and confirm it is a proper
distribution on a fine grid. `get_prior(fit)` returns paired `support` and
`density` vectors — the grid over the r-scale and the mass at each grid point —
plus summaries and metadata:

```{r verify-prior}
pr <- get_prior(fit)

sum(pr$density)              # a proper distribution: mass sums to 1
length(pr$support)           # a fine 1000-point grid over the r-scale
range(pr$support)            # the recovered support, on the r-scale
round(pr$mean, 3)            # the prior mean (r-scale)
pr$metadata$n_carriers       # how many units the prior was deconvolved from
```

The density sums to `r sum(get_prior(fit)$density)` — a proper distribution — over
a `r length(get_prior(fit)$support)`-point grid spanning roughly
$[`r paste(round(range(get_prior(fit)$support), 2), collapse = ",\\ ")`]$, with
prior mean `r round(get_prior(fit)$mean, 3)` on the r-scale. The prior was
deconvolved from `r get_prior(fit)$metadata$n_carriers` units (the 97 firms). Two
metadata fields are easy to misread: `pr$scale` is the literal label
`r dQuote(get_prior(fit)$scale, q = FALSE)` (a string naming the scale, not a
number), and `pr$metadata$alpha` is the log-spline coefficient **vector** (here of
length `r length(get_prior(fit)$metadata$alpha)`), the shape parameters of the
density, not a scalar penalty. Note what is *absent*: there is no Gaussian
"sigma" to read, because no Gaussian was assumed.

```{r fig-prior, fig.width = 7, fig.height = 4, fig.cap = "The deconvolved prior density G over the standardized r-scale, recovered by gradepath's native log-spline deconvolution with no parametric family assumed. The shaded area integrates to one; its spread is the estimated true spread of the latent effects, with the measurement noise removed.", fig.alt = "An area plot of the recovered prior density against the r-scale support running from about zero to one and a third. The density rises from zero, peaks near the prior mean of about a third, and tapers into a right tail, a unimodal but mildly right-skewed shape rather than a symmetric bell."}
ggplot(data.frame(v = pr$support, g = pr$density), aes(v, g)) +
  geom_area(fill = "steelblue", alpha = 0.7, colour = "steelblue4") +
  labs(x = "latent effect v (r-scale)", y = "prior density g(v)") +
  theme_gradepath()
```

The shaded shape is the recovered world: the spread of latent effects with the
measurement noise stripped out, narrower than the histogram of the raw $r_i$ would
be. That narrowing is the deconvolution doing its work — the apparent excess
spread in the raw estimates was noise, and the prior has set it aside.

## 3. The empirical-Bayes posterior

With $G$ in hand, each unit's posterior follows by Bayes' rule: combine the unit's
own likelihood — its estimate at noise scale $s_i$ — with the shared prior $\hat G$
[@walters2024empirical]:

$$
f\!\big(\theta_i \mid \hat\theta_i, s_i; \hat G\big)
\;=\;
\frac{\tfrac{1}{s_i}\,\phi\!\big((\hat\theta_i - \theta_i)/s_i\big)\,\hat g(\theta_i)}
     {\displaystyle\int \tfrac{1}{s_i}\,\phi\!\big((\hat\theta_i - x)/s_i\big)\,\hat g(x)\,dx} .
$$

The numerator is the product of two factors. The **likelihood** $\phi(\cdot)$
concentrates around the unit's own estimate $\hat\theta_i$; the **prior** $\hat g$
concentrates where units like this one usually land. The posterior is their
normalized product, and because $\hat g$ lives on a discrete grid the integral is a
finite sum — no simulation is needed in the one-level case.

The reported **posterior mean** and **posterior standard deviation** are the first
two moments of this distribution, and the credible interval
$[\text{lower},\,\text{upper}]$ is read off its percentiles at the fit's interval
level. The empirical-Bayes signature is **shrinkage**, with a precise shape: a
unit measured tightly (small $s_i$) has a sharp likelihood that dominates the
prior, so its posterior mean sits near its estimate; a unit measured loosely (large
$s_i$) has a flat likelihood the prior overrides, so its posterior mean is pulled
hard toward the body of $G$. **The noisier the estimate, the harder it shrinks** —
the empirical-Bayes generalization of James–Stein shrinkage to a nonparametric
prior. The credible interval is correspondingly *narrower* than a naive
$\pm$ standard-error band, because the prior contributes information.

**Verification panel.** Shrinkage leaves two checkable fingerprints. First, the
posterior means are *less spread* than the raw estimates — pulling everything
toward a common prior compresses the range. Second, the posterior standard
deviations are *smaller* than the raw standard errors — borrowing strength sharpens
nearly every unit (96 of 97). `as.data.frame(get_posterior(fit))` lays the units out one per row,
with the raw `estimate`/`se` alongside `posterior_mean`/`posterior_sd` and the
interval endpoints:

```{r verify-posterior}
post <- get_posterior(fit)
pdf  <- as.data.frame(post)

names(pdf)                          # raw (estimate, se) and posterior columns

# fingerprint 1: the posterior means are pulled in (less spread than the estimates)
c(estimate_sd       = sd(pdf$estimate),
  posterior_mean_sd = sd(pdf$posterior_mean))

# fingerprint 2: the posterior uncertainty is smaller than the raw se
c(mean_se           = mean(pdf$se),
  mean_posterior_sd = mean(pdf$posterior_sd))
```

Both fingerprints hold. The raw estimates spread with
$\mathrm{sd} = `r round(sd(as.data.frame(get_posterior(fit))$estimate), 3)`$, while
the posterior means spread with the smaller
$`r round(sd(as.data.frame(get_posterior(fit))$posterior_mean), 3)`$ — the means
are pulled in. The average raw standard error
$`r round(mean(as.data.frame(get_posterior(fit))$se), 3)`$ exceeds the average
posterior standard deviation
$`r round(mean(as.data.frame(get_posterior(fit))$posterior_sd), 3)`$ — nearly every unit's
uncertainty has shrunk (96 of 97; one unit, krw_race_062, edges up where the prior is more diffuse). Neither number was assumed; both fall out of combining the
estimates with the deconvolved $G$.

```{r fig-shrinkage, fig.width = 6.5, fig.height = 5, fig.cap = "Shrinkage made visible: each unit's posterior mean (vertical) against its raw estimate (horizontal), with the 45-degree line of no shrinkage. Points fall inside the line, pulled toward the prior, and the most extreme estimates are pulled the most.", fig.alt = "A scatter plot of posterior mean against raw estimate for 97 units, with a dashed 45-degree reference line. The cloud sits inside the line and is flatter than it, so the largest estimates have posterior means well below them and the smallest estimates have posterior means above them, the signature of shrinkage toward a common prior."}
ggplot(pdf, aes(estimate, posterior_mean)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.7, colour = "steelblue4") +
  labs(x = "raw estimate (r-scale)", y = "posterior mean (r-scale)") +
  theme_gradepath()
```

The points sit *inside* the 45-degree line and are flatter than it — shrinkage
drawn in two dimensions. The largest raw estimates, which are also the noisiest,
have posterior means well below the line; the units near the prior mean barely
move.

## 4. The pairwise outranking matrix Pi

The grades of m4 read neither the prior nor a posterior mean directly. They read a
single matrix that summarizes every pairwise comparison. For each ordered pair
$(i, j)$,

$$
\pi_{ij} \;=\; \Pr\!\big(\theta_i > \theta_j \mid \text{data}\big)
\;=\; \iint \mathbf{1}(\theta_i > \theta_j)\;
f(\theta_i \mid \hat\theta_i)\, f(\theta_j \mid \hat\theta_j)\, d\theta_i\, d\theta_j ,
$$

the posterior probability that unit $i$'s latent effect exceeds unit $j$'s
[@walters2024empirical]. The product form encodes that the two posteriors are
conditionally independent given the estimated prior $\hat G$, which holds in the
one-level model because $\hat G$ is treated as known. gradepath computes the
integral by CDF integration over the r-scale posterior weights, back-transforming
each unit's support to its own reporting scale before the comparison (the precision
adjustment is unit-specific), then stabilizes the result. The magnitude of
$\pi_{ij}$ is exactly *how confidently* the evidence orders one unit above another:
near $1$ or $0$ when the two posteriors barely overlap, near $0.5$ when they
coincide and the data cannot separate the units.

Two structural facts are enforced as deliberate invariants. A unit compared to
**itself** is a coin flip, so the diagonal is fixed at $0.5$. And a pair and its
mirror agree: $\pi_{ij} + \pi_{ji} = 1$, since ties are measure-zero under the
continuous prior — the lower triangle is filled as $1 - \pi_{ij}$ rather than
recomputed, so **antisymmetry** holds to machine zero. This matrix is the *only*
numeric object the grade integer program of m4 consumes; everything upstream of it
is empirical Bayes, everything downstream is social choice.

**Verification panel.** Pull the matrix and check all three structural facts: its
shape, its coin-flip diagonal, and its antisymmetry — both for a single pair and as
the worst-case deviation across every off-diagonal pair.

```{r verify-pairwise}
M <- get_pairwise(fit)$matrix

dim(M)                              # 97 x 97: one row and column per unit
unique(round(diag(M), 3))           # the diagonal: a unit vs itself is 0.5
M[1, 2] + M[2, 1]                   # one pair: exactly 1 (antisymmetry)
max(abs((M + t(M))[upper.tri(M)] - 1))   # worst-case antisymmetry deviation
```

The matrix is `r nrow(get_pairwise(fit)$matrix)` $\times$
`r ncol(get_pairwise(fit)$matrix)`, its diagonal is the single value
`r unique(round(diag(get_pairwise(fit)$matrix), 3))` (every unit ties itself), and
the first pair sums to
`r get_pairwise(fit)$matrix[1, 2] + get_pairwise(fit)$matrix[2, 1]`. The worst-case
antisymmetry deviation across all off-diagonal pairs is
`r max(abs((get_pairwise(fit)$matrix + t(get_pairwise(fit)$matrix))[upper.tri(get_pairwise(fit)$matrix)] - 1))`,
so $\pi_{ij} + \pi_{ji} = 1$ holds to machine zero. These checks confirm the matrix
carries the structure the grade program relies on.

```{r fig-contrast, fig.width = 7, fig.height = 6, fig.cap = "The grade-sorted pairwise outranking matrix Pi (KRW Figure 5a). Each cell is the posterior probability that the row unit's latent effect exceeds the column unit's; units are ordered by grade, so decisive comparisons (near zero or one) sit off the diagonal and indecisive ones (near one-half) cluster along it.", fig.alt = "A square heatmap of the 97-by-97 pairwise outranking matrix with units sorted by grade. The upper triangle is dark, indicating probabilities near one, and the lower triangle is light, near zero, with a band of intermediate shading along the diagonal where neighbouring units cannot be decisively ordered."}
gp_plot_posterior_contrast(fit)
```

Read the heatmap as a map of confidence. Far from the diagonal, where units of
very different grades meet, the cells are saturated — the posteriors barely
overlap, so the comparison is decisive. Near the diagonal, where neighbouring units
meet, the cells fade toward the middle tone of $0.5$ — those units cannot be
separated, which is exactly the information the grading must respect when deciding
who may share a grade.

## 5. The two-level extension (xi (x) eta)

Everything above is the **one-level** model: a single prior $G$ deconvolved from
the pooled ensemble. KRW's industry model adds one layer, and it is the same
machinery at a second scale. Split each firm's latent component into a
**within-industry** part $\xi_i$ (firm-specific, drawn once per firm) and a
**between-industry** part $\eta_{k(i)}$ (one industry effect per industry, shared
by every firm in it). For race the two combine multiplicatively,
$v_i = \eta_{k(i)}\,\xi_i$; for gender additively, $v_i = \eta_{k(i)} + \xi_i$.

KRW then deconvolve a **two-level prior** — the tensor product $G_\xi \otimes
G_\eta$ of a within-industry density and a between-industry density, each its own
log-spline on its own grid, fit jointly on a two-dimensional grid — and **push it
forward** to the $\theta$-scale over both supports at once. gradepath performs
this natively, as an internal theta-pushforward step (`gp_pushforward_theta`, an
internal kernel the two-level pipeline calls — not a verb you call yourself). Two
features have no one-level analogue.
The **pushforward** marginalizes the joint prior over *both* $\xi$ and $\eta$,
which the single-support one-level transform cannot express. And the pairwise
integral picks up a **same-industry override**: because two firms in one industry
share the single effect $\eta_k$, their conditional independence (§4) fails, and
the joint importance weight counts the shared industry likelihood **once**,
$L_{ij} = L_i$, rather than as the product $L_i\,L_j$ — using the product would
square that one piece of shared industry evidence and bias every same-industry
entry of $\Pi$.

Neither feature is exercised by the fit loaded here, because **this fit is
one-level** — there is no industry structure in it to verify against. So the
two-level deconvolution, its pushforward, and the same-industry override are
derived and verified in `r vlink("m5-two-level-and-calibration", "m5")` (the M2
surface), against the industry fixtures. Do not reach for an industry fit here; the
one-level objects of §2–§4 are the whole story for this vignette. The point to
carry forward is only that the prior → posterior → $\Pi$ arc extends to industries
by adding the between-industry layer and the override, with everything downstream
of $\Pi$ unchanged.

## 6. Where to next

With the prior deconvolved, the posteriors shrunk, and the pairwise matrix checked
for its structure, the EB core is complete and the pipeline turns to grading:

- **Foundations and notation** — `r vlink("m1-foundations-and-notation", "m1")`
  fixes every symbol used above — $G$, $\pi_{ij}$, the r-scale — and the two frozen
  invariants the later stages assume.
- **Precision and standardization** —
  `r vlink("m2-precision-and-standardization", "m2")` derives the r-scale and the
  carriers $\hat m$, $V_m$ that this vignette's deconvolution consumes; its §7 is
  the one-paragraph version of what m3 received.
- **Grading frontier and report cards** —
  `vignette("m4-grading-frontier-and-report-cards")` feeds the pairwise matrix
  $\pi_{ij}$ of §4 to the grade integer program, producing grades that are integer
  labels in $\{1, \dots, n\}$ — not letter grades — along the $\lambda$ frontier.
- **Two-level and calibration** —
  `vignette("m5-two-level-and-calibration")` builds and verifies the two-level
  industry model of §5 — the $\xi \otimes \eta$ deconvolution, the pushforward, and
  the same-industry override — and validates the whole pipeline by Monte Carlo.

For the workflow side, `r vlink("a1-getting-started", "a1")` is the first contact,
`r vlink("a2-the-grading-workflow", "a2")` runs this same deconvolution and
posterior stage hands-on with race and gender side by side,
`r vlink("a3-reading-and-exporting-results", "a3")` covers the posterior and
pairwise accessors used above, `r vlink("a4-figures", "a4")` is the figure gallery
the prior and contrast plots come from, and
`r vlink("a5-solvers-and-calibration", "a5")` covers the calibration harness.
Throughout, gradepath's printed output is result-first and quantity-anchored (the
package decision GP-DEC-14-A): it names the quantity — a posterior gap, an
outranking probability — rather than a verdict, so the same code path serves both
demographics without the printed text inverting on one of them.

### Provenance

```{r provenance}
sessionInfo()
```

## References
