---
title: "M4: Grading, the frontier, and report cards"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This vignette is the social-choice climax of the method track: how the pairwise
  outranking matrix Pi that m3 hands forward becomes grades. gradepath solves a
  Bayes-risk integer program that sorts the units into a small number of
  contiguous grades, trading the discordance rate against the posterior Kendall
  agreement through a single coarseness weight lambda. Sweeping lambda traces the
  information-reliability frontier: fewer grades are coarser but more reliable,
  more grades are finer but less reliable, and the published grading sits at the
  baseline lambda = 0.25. The selected grading is assembled into a report card,
  ranked from the most-extreme unit, with a conditional-discordance matrix that
  certifies each between-grade cut. We derive the integer program, the frontier,
  and the report card, citing KRW Propositions 1-3. The grade solve takes minutes,
  so it is shown but never run here: every step is verified live against the
  already-solved 97-firm parity fit.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M4: Grading, the frontier, and report cards}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

This is the fourth method-track vignette and the social-choice climax of the
pipeline. `r vlink("m3-deconvolution-and-posterior", "m3")` ended by deriving the
one object the grading reads: the pairwise outranking matrix $\Pi = (\pi_{ij})$,
whose entry $\pi_{ij} = \Pr(\theta_i > \theta_j \mid \text{data})$ is the
posterior probability that unit $i$'s latent effect exceeds unit $j$'s. This
vignette takes $\Pi$ and derives the three objects that finish the KRW pipeline:
the **Bayes-risk grade integer program** that sorts the units into contiguous
grades (§2-§3), the **information-reliability frontier** swept over the coarseness
weight $\lambda$ (§4), and the **report card** with its conditional-discordance
matrix (§5-§6). We summarize the theory from Kline, Rose & Walters
[@kline2024discrimination] — stating, not reproving, their Propositions 1-3 (§7)
— and follow each step with a verification panel checked live against the fit.

We load that fit once, at the top. The grade integer program *solves*: it is a
mixed-integer program that takes minutes at $n = 97$, so the verbs `gp_grade()`
and `gp_grade_path()` appear only as `eval = FALSE` illustrations. Everything else
is a fast read off the **already-solved** path stored on the fit, so **no integer
program is solved anywhere in this document**.

```{r load-fit}
fit <- gp_parity_fit("race")   # the proven 97-firm RACE card, grades 2 / 81 / 14
```

This is the 97-firm RACE parity fit at the KRW baseline $\lambda = 0.25$, already
solved across the $\lambda$ grid. Keep one fact from
`r vlink("m1-foundations-and-notation", "m1")` in view: a grade is an **integer
label** in $\{1, \dots, n\}$ — not a letter grade such as A or B — and exported
grade 1 is the most-extreme block in $\theta$. A grade is a sorting by posterior
evidence, never a contest.

## 1. From the pairwise matrix to grades

m3 handed us $\Pi$, a full $n \times n$ table of pairwise outranking
probabilities. Why is that not already the answer? Because $\Pi$ orders every
*pair* but not the *whole*. The naive rule — separate $i$ above $j$ whenever
$\pi_{ij}$ clears a threshold — can disagree with itself: the evidence might
decisively order $i$ above $k$ yet be too weak to order either against a middle
unit $j$, leaving the pairwise rule to tie $i$ with $j$ and $j$ with $k$ while
demanding $i$ above $k$. That is a cycle, the report-card analog of a Condorcet
cycle in voting, and a grading cannot contain one.

So the goal is precise: partition the $n$ units into a small number of
**contiguous, ordered grades** so that the ordering respects the posterior where
the evidence is strong, pools units the data cannot separate, and is cycle-free.
Because the conflicts must be resolved *globally* rather than pair by pair, this
is an optimization, not a thresholding pass. gradepath follows KRW
[@kline2024discrimination] in casting it as a Bayes-risk integer program — the
modern compound-decision view of ranking and selection [@gukoenker2023;
@mogstad2024]. The arc below states the program (§2), extracts the grade from its
solution (§3), sweeps $\lambda$ for the frontier (§4), assembles the report card
(§5), decomposes its reliability boundary by boundary (§6), and places the result
against KRW's optimality propositions (§7).

## 2. The Bayes-risk integer program

The grading is encoded by binary pairwise indicators. For each *ordered* pair
$(i, j)$ with $i \neq j$, the variable $D_{ij} \in \{0, 1\}$ is set to $1$ exactly
when the grading ranks $i$ above $j$ (so $i$ carries the more-extreme grade).
Crucially, $D_{ij}$ and $D_{ji}$ are **independent** solver variables — the
program does *not* hard-impose $D_{ij} + D_{ji} = 1$; antisymmetry and
transitivity enter as linear constraints, which is exactly what lets two units
*tie* when the evidence is weak.

**The objective.** The solver minimizes a Bayes risk that is linear in the
$D_{ij}$. For each ordered pair the per-pair contribution of separating $i$ above
$j$ is

$$
\Pi_{ji} \;-\; \lambda\, \Pi_{ij}
\qquad\bigl(\text{equivalently } P[j,i] - \lambda\,P[i,j]\bigr),
$$

summed over the ordered pairs the grading separates, where $\Pi = (\pi_{ij})$ is
the pairwise matrix from m3. Read the two terms as a ledger. The first,
$\Pi_{ji} = \pi_{ji}$, is the **discordance cost**: the posterior mass saying $j$
is the more-extreme unit, charged against a grading that put $i$ above it. The
second, $-\lambda\,\Pi_{ij} = -\lambda\,\pi_{ij}$, is the **information reward**,
scaled by $\lambda$: the agreeing mass, credited for a separation the evidence
supports. The optimizer is paid to order a pair the way the posterior does and
penalized for the reverse; $\lambda$ is the exchange rate between the two.

**The constraints.** Two families pin down the feasible set. *Antisymmetry*
forbids ranking $i$ above $j$ and $j$ above $i$ at once; *transitivity* forbids
cycles, so $i$ above $j$ and $j$ above $k$ force $i$ above $k$. gradepath encodes
transitivity through an integer grade-label auxiliary (every unit gets a label,
the binaries tied to the label gaps), making cycles infeasible by construction;
the backend is the subject of `r vlink("a5-solvers-and-calibration", "a5")`. The
one constraint a reader should carry is the **separation threshold**: two units
may share a grade only when neither outranks the other decisively, i.e. when

$$
\max(\Pi_{ij},\, \Pi_{ji}) \;<\; \frac{1}{1+\lambda}.
$$

At the baseline $\lambda = 0.25$ the threshold is $1/1.25 = 0.80$: the stronger
pairwise probability must fall below $0.80$ for two units to pool. Raise $\lambda$
and the threshold drops toward $0.5$ (separate on the faintest majority); lower it
and the threshold rises toward $1$ (pool unless near-certain). This inequality is
the gear that turns $\lambda$ into a grade count.

**The call (shown, not run).** The grading is produced by `gp_grade_path()`, which
solves one integer program per $\lambda$ on the grid and records the reportable
$\lambda$; `gp_grade()` is the one-shot for a single $\lambda$. Both *solve* and
take minutes, so they are shown here only as illustrations:

```{r grade-call, eval = FALSE}
# Solve the full grading frontier from a pairwise object (the main verb).
# This SOLVES one integer program per lambda and takes minutes -- not run here.
path <- gp_grade_path(get_pairwise(fit),
                      lambda_grid     = c(0.25, 0.50, 1.00),  # 0.25 must be present
                      selected_lambda = 0.25)                 # the reportable weight

# One-shot: solve a single lambda and return its fit.
one  <- gp_grade(get_pairwise(fit), lambda = 0.25)
```

We never run these. The path on the bundled fit is already solved, and the solver
returns the optimum exactly — status `"optimal"` at a Gurobi MIP gap of $0$ —
which we read straight off the fit:

```{r verify-optimal}
fit$selected_grade$summary$status        # "optimal": the IP is solved to optimality
fit$selected_grade$summary$grade_count   # the selected grading has 3 grades
fit$selected_grade$summary$n_units       # over all 97 firms
```

The selected grading is solved to optimality (`status = `
`r dQuote(fit$selected_grade$summary$status, q = FALSE)`), with
`r fit$selected_grade$summary$grade_count` grades over all
`r fit$selected_grade$summary$n_units` firms. Because the gap is $0$, this is the
*proven* risk-minimizer at $\lambda = 0.25$, not an approximation. (Read the
scalar objective from `fit$grade_path$summary$objective`, shown in §4 — *not* from
`fit$selected_grade$objective`, which is a structured list rather than a number.)
The optimality guarantees are KRW Propositions 1-3, stated in §7.

## 3. Extracting the grade

The integer-label auxiliary that enforced transitivity is **not** the reported
grade. The grade is reconstructed from the optimal binaries by a win-count and a
dense rank — the grade-direction triple fixed in
`r vlink("m1-foundations-and-notation", "m1")`. First, for each unit sum the
binaries to get its **win-count**,

$$
S_i \;=\; \sum_{j} D_{ij},
$$

the number of units $i$ is placed above (the code uses $D_{ij}$, not $D_{ji}$,
because the two are independent). Second, the exported grade is the **descending
dense rank** of the win-count,

$$
\text{grade}_i \;=\; \operatorname{dense\_rank}\!\big(S_i;\ \text{ascending}=\text{FALSE}\big),
$$

so the unit with the **largest** $S_i$ — placed above the most others, the
**most-extreme $\theta$** — earns the smallest label, grade $\mathbf{1}$, and
equal win-counts share a grade. The welfare reading of grade 1 *flips* between
datasets: for firms it is the most-discriminatory firm, for names the
best-treated name (the m1 firms-vs-names invariant). The machinery is identical;
only the interpretation reverses — which is exactly why gradepath's output is
result-first and quantity-anchored (the package decision `GP-DEC-14-A`).

**Verification panel.** The internal $D_{ij}$ and $S_i$ are solver-internal, but
the assembled card carries their *consequence*: grade 1 sits first, the
largest-$S_i$ block at the top. `as.data.frame(fit)` lays the 97 firms out one per
row in rank order:

```{r verify-card-head}
card <- as.data.frame(fit)
dim(card)                       # 97 rows x 10 columns: one firm per row
head(card, 6)                   # grade 1 first -- the dense-rank extraction
```

The card is `r nrow(as.data.frame(fit))` $\times$ `r ncol(as.data.frame(fit))` and
its first rows carry grade `r head(as.data.frame(fit)$grade, 1)`, consistent with
the descending dense-rank extraction placing the largest-$S_i$ block first. The directional consequence is
also checkable numerically: if grade 1 is the most-extreme $\theta$, the mean
posterior effect must be **largest** in grade 1 and decrease as the grade number
rises. Group the card's `posterior_mean` by grade:

```{r verify-direction}
tapply(card$posterior_mean, card$grade, mean)   # decreasing across grades 1 -> 3
```

Grade 1 carries the highest mean posterior effect
(`r round(tapply(as.data.frame(fit)$posterior_mean, as.data.frame(fit)$grade, mean)[["1"]], 3)`),
falling to
`r round(tapply(as.data.frame(fit)$posterior_mean, as.data.frame(fit)$grade, mean)[["2"]], 3)`
in grade 2 and
`r round(tapply(as.data.frame(fit)$posterior_mean, as.data.frame(fit)$grade, mean)[["3"]], 3)`
in grade 3 — monotonically decreasing, the numeric confirmation that the
dense-rank-of-win-count extraction places the most-extreme $\theta$ in grade 1.

## 4. The information-reliability frontier

A single $\lambda$ gives a single grading; sweeping $\lambda$ over a grid gives a
*path* of gradings, and scoring each against the same $\Pi$ traces the
**information-reliability frontier**. Each point is the risk-minimizer at its own
$\lambda$, so the whole curve is the menu of defensible gradings. The two
coordinates are the **reliability** $1 - \overline{\mathrm{DR}}$ on the horizontal
axis — the posterior share of between-grade comparisons that agree with the
posterior — and the posterior expected Kendall agreement $\bar\tau$ on the
vertical — how much rank information the grading carries. By construction,

$$
\mathrm{Reliability} \;=\; 1 - \overline{\mathrm{DR}}.
$$

The trade-off is monotone, and it is the central finding. As $\lambda$ rises the
optimizer separates more units, so the **grade count rises**, $\bar\tau$ rises
(more information), the discordance rate rises, and reliability $1 -
\overline{\mathrm{DR}}$ **falls**. Fewer grades are coarser but more reliable;
more grades are finer but less reliable. There is no free lunch: every grade you
add buys information at the cost of reliability. At $\lambda = 1$ the rule
collapses to the Condorcet ordering, the finest the rule will return.

**Verification panel.** The frontier table is `fit$grade_path$summary`, one row per
$\lambda$. For the parity fit it has three rows; read off the monotone
grade-count-up / reliability-down trade-off:

```{r verify-frontier}
fit$grade_path$summary[, c("lambda", "grade_count", "discordance_rate",
                           "reliability", "tau_bar")]
```

Three rows tell the whole story. As $\lambda$ rises $0.25 \to 0.50 \to 1.00$, the
grade count climbs `3 -> 6 -> 97`, the discordance rate rises
`0.039 -> 0.161 -> 0.268`, reliability — its complement — *falls*
`0.961 -> 0.839 -> 0.732`, and $\bar\tau$ rises `0.207 -> 0.428 -> 0.464`. The
endpoint $\lambda = 1$ has shattered into
`r fit$grade_path$summary$grade_count[fit$grade_path$summary$lambda == 1]`
singleton grades — the Condorcet ordering — while the selected grading sits at the
other end, $\lambda = 0.25$, with
`r fit$grade_path$summary$grade_count[fit$grade_path$summary$lambda == 0.25]`
grades. The objective at the selected $\lambda$ is the optimal Bayes risk in the
program's coordinates:

```{r verify-objective}
fit$grade_path$summary$objective[fit$grade_path$summary$lambda == 0.25]
```

— about
`r round(fit$grade_path$summary$objective[fit$grade_path$summary$lambda == 0.25], 1)`,
the minimized risk of the selected grading (more negative is better, since the
information reward enters with a minus sign), with every row reported `optimal`.

**Figure.** `gp_frontier()` scores the already-solved path against $\Pi$ and
`gp_plot_frontier()` draws the $(1 - \overline{\mathrm{DR}}, \bar\tau)$ curve (KRW
Figure 6b) — a fast read, no solve. The two-row legend sits at the bottom, so give
the figure a wide, short canvas:

```{r fig-frontier, fig.width = 8, fig.height = 5.5, fig.cap = "The information-reliability frontier for the RACE fit (KRW Figure 6b): reliability (1 minus the discordance rate, horizontal) against the posterior expected agreement tau-bar (vertical), with the selected lambda = 0.25 grading marked. Walking the curve trades grade count against reliability.", fig.alt = "A scatter-and-line plot. The horizontal axis is reliability, one minus the discordance rate; the vertical axis is the posterior expected agreement tau-bar. A connected path of points traces the frontier from few grades at high reliability to many grades at lower reliability, with the selected grading highlighted and a two-row legend at the bottom."}
fr <- gp_frontier(fit$grade_path, pairwise = get_pairwise(fit))
gp_plot_frontier(fr)
```

Read the curve from right to left as a walk in $\lambda$. The right-most,
most-reliable point is the coarse three-grade baseline; moving left adds grades,
buying $\bar\tau$ at the cost of reliability. The highlighted point is the selected
grading — a *small* number of grades capturing most of the achievable $\bar\tau$
while holding the discordance rate near
`r round(fit$grade_path$summary$discordance_rate[fit$grade_path$summary$lambda == 0.25], 3)`.

## 5. The report card

The report card translates the frontier's abstract trade-off into the policy
artifact a regulator reads: a ranked list of units, each with its posterior
interval and its selected grade. Its assembly is a join with **two keys**. The
**sort key** is the Condorcet ($\lambda = 1$) order — the finest ranking, the
most-extreme unit on top — so the rows read top-to-bottom as most-to-least
extreme. The **display key** is the selected ($\lambda = 0.25$) grade — the small
number of statistically defensible bands painted onto that fine ranking. One row
order, one grade column, no recomputation.

The assembled object is `get_report_card(fit)`, and its tidy form is
`as.data.frame(fit)` — the `r ncol(as.data.frame(fit))`-column card seen in §3,
carrying `id`, `label`, `grade`, `sort_rank`, `selected_lambda`, `posterior_mean`,
the interval `lower`/`upper`, and the raw `estimate`/`se`. For the race fit, grade
1 is the small most-discriminatory block, grade 3 the largest best-behaved block.

**Figure.** `gp_plot_report_card()` draws the point-and-interval "caterpillar"
(KRW Figure 7), one firm per row, the most-extreme on top, coloured by grade:

```{r fig-report-card, fig.width = 7, fig.height = 8, fig.cap = "The RACE report card (KRW Figure 7): each of the 97 firms as its posterior White-minus-Black contact-rate gap (point) with its credible interval (horizontal line), one firm per row, ranked with the most-extreme on top and coloured by its integer grade in the 2 / 81 / 14 split.", fig.alt = "A horizontal caterpillar plot of 97 firms, one per row, ranked from most extreme at the top to least extreme at the bottom. Each row shows a point for the firm's posterior gap and a horizontal line for its credible interval. Points are coloured by grade into three groups, with the colour changing where the intervals stop overlapping."}
gp_plot_report_card(get_report_card(fit))
```

Read it in three moves: vertical position is rank (top = most extreme), horizontal
position is the shrunken posterior gap, and line length is uncertainty. The grade
colours change exactly where the intervals stop overlapping — the bands are cut by
the evidence, not at arbitrary points. The card is a *ranking* annotated with the
bands the data can defend, never a verdict.

## 6. Conditional discordance

The headline discordance rate of §4 is an average, and an average can hide one
unreliable boundary. The **conditional discordance matrix** decomposes it grade by
grade. For an ordered pair of grades $(g, g')$ with $g$ the more extreme, the cell
$\mathrm{DR}_{g,g'}$ is the share of the between-block posterior mass that is
*misranked* — the probability that a unit drawn from grade $g$ is actually less
extreme than one drawn from grade $g'$, conditional on the pair straddling that
boundary. A small off-diagonal cell certifies that *that specific cut* is clean,
and it reads as posterior odds: a discordance of $\mathrm{DR}_{g,g'} = 0.05$ leaves
$1 - 0.05 = 0.95$ posterior mass on the correct ordering, i.e. $19{:}1$ posterior
odds that the grade ordering is right.

**Figure.** `gp_plot_discordance()` reads the same `fr` object and draws the
lower-triangular matrix (KRW Figure 10a) — no solve:

```{r fig-discordance, fig.width = 6, fig.height = 5, fig.cap = "The conditional discordance matrix for the RACE fit (KRW Figure 10a): a lower-triangular heatmap of the between-grade discordance rate for each pair of grades, the more-extreme grade on the columns (grade 1 at the left) and the less-extreme grade on the rows, each cell annotated with its rate. Small off-diagonal values mean clean cuts between grades.", fig.alt = "A lower-triangular heatmap with one cell per pair of grades. Columns run from the most-extreme grade on the left; rows run from the second grade at the top down. Each cell is shaded by its between-grade discordance rate, darker for higher discordance, with the numeric rate written inside; the off-diagonal cells are light, indicating clean separation."}
gp_plot_discordance(fr)
```

The off-diagonal cells stay small: every between-grade cut in the three-grade
baseline is reliable, and no single boundary carries a disproportionate share of
the misranking. This is the granular complement to the headline scalar — the
aggregate discordance rate of
`r round(fit$grade_path$summary$discordance_rate[fit$grade_path$summary$lambda == 0.25], 3)`
is spread evenly across the cuts, not concentrated in one fragile boundary.

## 7. Propositions 1-3

The Bayes-risk program is not an ad-hoc rule. KRW prove that its optimum inherits
the axiomatic structure of well-studied voting rules, with the classical $0.5$
majority threshold sharpened to $1/(1+\lambda)$. We **state** their Propositions
1-3 and cite the paper [@kline2024discrimination]; the proofs are in their Online
Appendix B and are not reproduced here. The three are a hierarchical guarantee — a
dominant individual, a dominant set, the internal structure of that set:

- **Proposition 1 ($\lambda$-Condorcet criterion).** If a unit $i$ satisfies
  $\pi_{ij} > 1/(1+\lambda)$ for every other $j$, then $i$ receives a strictly
  more-extreme grade than every other unit — equivalently the largest win-count
  $S_i$, exported grade 1. At $\lambda = 1$ the threshold is the classical
  Condorcet value $0.5$; at the baseline $\lambda = 0.25$ it is $0.80$. A
  second-place analogue holds in turn: if a further unit $k$ satisfies
  $\pi_{kj} > 1/(1+\lambda)$ for every remaining $j \neq i$, then $i$ receives a
  more-extreme grade than $k$, which in turn receives a more-extreme grade than
  all the remaining units — the criterion extends to the runner-up.
- **Proposition 2 ($\lambda$-Smith criterion).** If a *set* $S$ satisfies
  $\pi_{ij} > 1/(1+\lambda)$ for all $i \in S$ and $j \notin S$, then every unit
  receiving the most-extreme grade lies in $S$. The grading respects dominant
  coalitions, not just dominant individuals.
- **Proposition 3 (unordered $\lambda$-Smith candidates).** Under Proposition 2,
  if additionally $\pi_{ij} < 1/(1+\lambda)$ for all $i, j \in S$, then every unit
  in $S$ receives the **same** grade: the set collectively separates from
  outsiders, but the within-$S$ evidence is too weak to separate its members.

Together these say the $\lambda$-optimal grading treats *ties as a first-class
feature* rather than an edge case — exactly the honesty the report card needs, and
the guarantee behind the `optimal` status verified in §2. The modern reading of
ranking-and-selection as a compound decision [@gukoenker2023; @mogstad2024] places
these results in the same lineage as the empirical-Bayes shrinkage of m3.

## 8. Where to next

With the integer program derived, the grade extracted, the frontier swept, and the
report card and its discordance matrix assembled and verified live on the fit, the
KRW pipeline is complete — the pairwise matrix of m3 has become a small number of
honest grades. The next steps:

- **Foundations and notation** — `r vlink("m1-foundations-and-notation", "m1")`
  fixes the grade-direction triple ($S_i = \sum_j D_{ij}$, dense rank descending,
  grade 1 = most-extreme $\theta$) and the $1/(1+\lambda)$ threshold this vignette
  builds on.
- **Precision and standardization** —
  `r vlink("m2-precision-and-standardization", "m2")` derives the r-scale and the
  $\beta$-GMM step upstream of the pairwise matrix.
- **Deconvolution and posterior** —
  `r vlink("m3-deconvolution-and-posterior", "m3")` builds the prior, the
  posteriors, and the pairwise matrix $\Pi$ that this vignette's integer program
  consumes.
- **Two-level and calibration** —
  `r vlink("m5-two-level-and-calibration", "m5")` extends the grading to the
  two-level industry model and validates the pipeline by Monte Carlo.

For the workflow side, `r vlink("a1-getting-started", "a1")` is the first contact,
`r vlink("a2-the-grading-workflow", "a2")` runs the full grading stage hands-on
with race and gender side by side,
`r vlink("a3-reading-and-exporting-results", "a3")` covers the report-card and
frontier accessors used above, `r vlink("a4-figures", "a4")` is the figure gallery
the three plots come from, and
`r vlink("a5-solvers-and-calibration", "a5")` covers the solver backends that run
the integer program this vignette only shows. Throughout, the grade is a sorting
by posterior evidence, never a contest: gradepath's printed output is result-first
and quantity-anchored (the package decision `GP-DEC-14-A`), so the same code path
serves firms and names without the printed text inverting on one of them.

### Provenance

```{r provenance}
sessionInfo()
```

## References
