---
title: "M5: Two-level industries and calibration"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This is the last method-track vignette, and it has two parts. The first
  extends the one-level grading to industries. The industry model splits each
  firm's standardized effect into a within-industry component and a
  between-industry component, deconvolves the two as a tensor-product prior,
  and pushes that prior forward to the firm scale with gradepath's native
  theta-pushforward. Two firms in the same industry share one between-industry
  effect, so it cancels in their comparison and they are compared on the
  within-industry component alone. The industry pipeline grades a firm-level
  route and a between-industry route and assembles two report cards. We report
  its status honestly as PARTIAL_ACCEPTED: the exact industry grade counts pass,
  but the continuous discordance, agreement, and fit rows are classified by
  fixture parity, not by direct reproduction of the paper's industry values, and
  we read that ledger live. The second part covers gp_calibrate, the seeded
  Monte-Carlo that checks the grading method's frequentist guarantees on a
  fitted model, and explains why KRW's Table E1 study is deliberately deferred.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M5: Two-level industries and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
{ p <- system.file("scripts/_setup.R", package = "gradepath"); if (!nzchar(p)) p <- "_setup.R"; source(p) }
```

This is the fifth and final method-track vignette. The first four built and
verified the one-level pipeline end to end: `r vlink("m1-foundations-and-notation", "m1")`
fixed the notation, `r vlink("m2-precision-and-standardization", "m2")` derived
the precision exponent and the r-scale,
`r vlink("m3-deconvolution-and-posterior", "m3")` built the prior, the
posteriors, and the pairwise matrix $\Pi$, and
`r vlink("m4-grading-frontier-and-report-cards", "m4")` turned $\Pi$ into grades
along the information-reliability frontier. This vignette closes the track with
two extensions that sit *around* that core rather than inside it.

The first half is the **two-level industry model** — the same grading machinery,
but applied to firms nested in industries, with a between-industry effect that
every firm in an industry shares. The second half is **calibration** — a seeded
Monte-Carlo, `gp_calibrate()`, that checks whether the grading method's
frequentist guarantees survive estimating the prior. Both halves summarize the
theory and cite Kline, Rose & Walters [@kline2024discrimination]; the formal
two-level statements are in their Online Appendix and the Monte-Carlo design is
their Table E1 study.

Two disciplines run through the vignette. **Nothing solves here:** the two-level
grade solve and `gp_calibrate()` each run an integer program (the calibration runs
one *per draw*), so they appear only as `eval = FALSE` illustrations, and the one
live panel reads the M2 acceptance ledger, a fast lookup. And the **honesty
discipline** of the method track holds: the industry surface is reported exactly as
`PARTIAL_ACCEPTED`, never inflated, and `gp_calibrate()` is named for what it is —
a qualitative diagnostic, not a reproduction of any published number.

# Part I — The two-level industry model

## 1. From one level to two: why industries

The one-level model of m1–m4 grades $n$ firms against one prior $G$ over a single
latent quantity. KRW's industry analysis asks a structured question on top of
that: *how much of a firm's measured discrimination is the firm, and how much is
its industry?* A staffing firm and a bank may both post a wide White-minus-Black
contact-rate gap, but if staffing as a whole runs wide while this particular bank
is wide *for a bank*, those two firms are extreme for different reasons. The
one-level model cannot see that distinction — it has one latent effect per firm
and no notion of the group the firm sits in.

The two-level model adds exactly that notion. It writes each firm's standardized
effect as the combination of a **within-industry** component, specific to the
firm, and a **between-industry** component, shared by every firm in the same
industry. With the effect decomposed this way, gradepath can grade firms *within*
the structure (the firm-level route) and grade the industries *themselves* (the
between-industry route), producing two report cards instead of one. Everything
downstream — the pairwise matrix, the grade integer program, the frontier — is
the same machinery m4 already derived; only the prior and the posterior that feed
it change. This is Milestone **M2**, deferred behind the one-level **M1** of the
earlier vignettes, and the honesty section (§5) is about exactly how far that
milestone got.

## 2. The two-level model: within, between, and the pushforward

We summarize the model; the full derivation is the package's design (its two-level
derivation), and the formal statements are KRW [@kline2024discrimination]. Write
$k(i)$ for the industry of firm $i$. The two-level decomposition of the firm's
precision-independent effect $v_i$ has a **within-industry** component $\xi_i$
(one per firm, drawn from a within-industry distribution $G_\xi$) and a
**between-industry** component $\eta_{k(i)}$ (one per industry, drawn from a
between-industry distribution $G_\eta$ and *shared* by every firm in industry
$k$). For the race panel the two combine multiplicatively,

$$
v_i \;=\; \eta_{k(i)}\,\xi_i,
\qquad
\theta_i \;=\; s_i^{\beta}\,\eta_{k(i)}\,\xi_i,
$$

while for the gender panel they combine additively,
$v_i = \eta_{k(i)} + \xi_i$, with $\theta_i = \mu + s_i^{\beta}(\eta_{k(i)} + \xi_i)$.
The familiar $s_i^{\beta}$ rescaling — the precision dependence of
`r vlink("m2-precision-and-standardization", "m2")` — is unchanged; what is new
is the split of the latent effect into a firm part and an industry part.

**The tensor-product prior.** The two components are deconvolved *jointly* into
the tensor product

$$
G \;=\; G_\xi \otimes G_\eta,
$$

a two-dimensional distribution on the $(\xi, \eta)$ grid: $G_\xi$ on the
within-industry support, $G_\eta$ on the between-industry support. gradepath fits
both at once by a penalized log-spline maximum likelihood — the native two-level
deconvolution, which reuses the one-level spline primitives but is not the
one-level engine, since the joint fit, its cluster-robust GMM moments, and its
support caps have no one-level analogue. A *within-share* constraint pins the
between component to the multiplicative or additive identity — $E[\eta] = 1$ for
race, $E[\eta] = 0$ for gender — so the industry layer carries only the shared
*departure* from the average firm, and the within component carries the rest.

**The theta-pushforward.** The prior lives on the $(\xi, \eta)$ scale, but the
grading reads a prior on the firm $\theta$-scale. The bridge is an internal
theta-pushforward step (gradepath's `gp_pushforward_theta`, an internal kernel
mirroring KRW's `get_g_theta.m`; the two-level pipeline runs it for you): it
marginalizes the joint prior over *both* supports, forming
$\theta = s^{\beta}\,\xi\,\eta$ (race) or $\theta = \mu + s^{\beta}(\xi + \eta)$
(gender) on a fixed grid and accumulating the joint mass $g_\xi \cdot g_\eta$ at
each $\theta$ value. The one-level pushforward of m3 takes a single support and
cannot do this — it has the wrong arity for two dimensions — so the two-level
pushforward is a separate native kernel. The output is an ordinary
$\theta$-scale prior, and from there the posterior, the pairwise matrix, and the
grade integer program are exactly the m3–m4 objects.

## 3. The same-industry override

The single most consequential — and most error-prone — line in the two-level model
is what happens when a pairwise comparison falls *within* an industry. The
posterior compares two firms by integrating an ordering indicator against the
firms' likelihood weights. When firm $i$ and firm $j$ sit in **different**
industries, their industry effects $\eta_{k(i)}$ and $\eta_{k(j)}$ are
independent draws, so the joint weight genuinely factorizes into the product of
the two firms' likelihoods,

$$
L_{ij} \;=\; L_i \cdot L_j
\qquad (k(i) \neq k(j)).
$$

When firm $i$ and firm $j$ sit in the **same** industry, they share the *one*
between-industry effect $\eta_k$. The likelihood evidence for that shared effect
has been observed once, not twice, so the joint weight must condition on the
single $\eta_k$ and the likelihood enters **once**:

$$
L_{ij} \;=\; L_i \;\bigl(= L_j\bigr)
\qquad (k(i) = k(j)).
$$

This is the **same-industry override**: within an industry the between-industry
component is common to both firms, so it *cancels* in their comparison, and the
within-industry component $\xi$ alone governs which firm is more extreme. Two banks
both inherit whatever the banking industry's effect is, so that shared inheritance
says nothing about which bank has the wider contact-rate gap — only their
firm-specific $\xi$ does.

Forming the product $L_i \cdot L_j = L_i^2$ for a same-industry pair would instead
*square* the industry's evidence, double-counting the one shared $\eta_k$ as though
it had been seen twice and biasing every within-industry comparison. The override
is scoped strictly to the two-level case — in the one-level model there is no shared
industry effect, and it must not fire — and it is social-choice plumbing in the
pairwise construction, not an empirical-Bayes step, so it propagates to every
pairwise matrix the two-level posterior produces.

## 4. The two-level pipeline

The industry analysis is three verbs, all of which need constructed two-level
inputs and a solve, so they are **shown but not run** here. The pipeline mirrors
the one-level path of m4 but carries two routes side by side: a firm-level route
(`industry_rfe`) graded from the firm $\Pi_\theta$, and a between-industry route
(`btwn`) graded from the industry-level $\Pi$ over the $\bar s\,\eta$ orderings.

**Build the two-level pairwise matrices.** `gp_twolevel_pairwise()` takes the
firm-level and between-industry pairwise inputs (with the same-industry override
already encoded in their integrals) and bundles them for the two routes:

```{r twolevel-pairwise, eval = FALSE}
# Builds the two-level outranking bundle for the two routes.
# Needs constructed two-level inputs -- shown, not run.
pw <- gp_twolevel_pairwise(Pi_theta, Pi_bar,
                           ids             = firm_ids,
                           industry_levels = industry_ids,
                           control         = gp_control(backend = "gurobi"))
```

**Solve the two-level grades.** `gp_twolevel_grade()` routes each bundle into the
package grade solver — one grade path per route, read off at the reporting
$\lambda$. It *solves*, so it is illustrative only:

```{r twolevel-grade, eval = FALSE}
# Solves one grade integer program per route -- shown, not run.
tlg <- gp_twolevel_grade(pw,
                         posterior = twolevel_posterior,  # backs the report cards
                         lambda    = 0.25)                 # the reporting penalty
```

The returned `gp_twolevel_grade` carries an `industry_rfe` slot (the firm-level
route: its grade path, selected grade, grade count, and — when a posterior is
supplied — a firm report card) and a `btwn` slot (the same payload for the
between-industry route), along with a combined `producer_status` that is `"OK"`
only when *both* routes are acceptance-ready.

**Select a report card.** `gp_twolevel_report_card()` returns the
posterior-backed card for the chosen route:

```{r twolevel-card, eval = FALSE}
# The firm-level industry card, or the between-industry card -- shown, not run.
firm_card <- gp_twolevel_report_card(tlg, model = "industry_rfe")
ind_card  <- gp_twolevel_report_card(tlg, model = "btwn")
```

One scope note the next section makes precise: a report card exists only when the
two-level grade was built *with* a posterior. While the industry surface is
`PARTIAL_ACCEPTED`, the package exports no public builder for a posterior-backed
two-level grade, so a grade assembled from bare $\Pi$ matrices is structural only
and carries no card — calling the selector on it raises an informative error that
points back to the status ledger. The verbs are documented and shown, but the
success path runs on internal fixtures, not on user-constructed input; that
deliberate boundary is itself part of the honest status.

## 5. The honest M2 status: PARTIAL_ACCEPTED

Here is the one live panel of the vignette, and the heart of the honesty story.
The two-level milestone did *not* reach full acceptance, and the package says so
in a structured ledger you can read instantly — no solve, just a lookup. The
public summary is `gp_m2_status()`:

```{r m2-status}
gp_m2_status()[, c("component", "status", "detail")]
```

Read the six rows plainly. They tell a precise, un-inflated story:

- **`formal` -> `PARTIAL_ACCEPTED`.** The overall verdict, with the reason
  `"ONE_OR_MORE_L02_ROWS_REMAIN_APPROXIMATE"`. One or more continuous rows did not
  clear the parity band, so the milestone is partial — not accepted, not failed.
- **`L01_industry_grade_counts` -> `PASS`.** The exact part. The industry grade
  **counts** are integer, optimizer-determined quantities, and they match the
  archive exactly (`"2/2 exact count rows pass"`). This is a genuine,
  registered, exact reproduction.
- **`L02_race_continuous` -> `APPROXIMATE_OK`.** The race continuous summaries
  (discordance, agreement, fit) remain *approximate*: the recorded $\Pi_\theta$
  fixture gap is `0.0121756787`, above the `0.01` promotion threshold. The status
  is honest — `APPROXIMATE_OK` means "documented as approximate," not "passed."
- **`L02_gender_continuous` -> `PROMOTED`.** The gender continuous summaries were
  *promoted to banded*: the recorded gender $\Pi_\theta$ fixture gap is
  `0.0066335820`, at or below `0.01`. Promotion means the gender panel's fixture
  artifacts met the intermediate parity band used for M2 — it does **not** mean
  the paper's industry discordance, agreement, or fit targets were reproduced
  directly. This is fixture-parity evidence, not paper-value reproduction.
- **`N10_support_guards` -> `EVIDENCE_OK`.** Four synthetic support guards carry
  evidence (`"4/4 N10 rows carry synthetic evidence status"`). These are guards
  on the support construction, not registered paper-value passes.
- **`M1_status_boundary` -> `NOT_RECALCULATED`.** The M2 ledger is M2-only. It
  reads the two-level surface and deliberately does *not* touch or recompute the
  one-level M1 acceptance contract.

The honest reading, stated once and plainly: **M2 is `PARTIAL_ACCEPTED`.** The
exact industry grade counts pass (L01). The continuous discordance, agreement,
and fit rows are classified by *fixture parity* — gender promoted to banded at gap
`0.0066`, race still approximate at gap `0.0122 > 0.01` — and a promotion is
intermediate-band evidence, **not** direct reproduction of the paper's industry
values. The N10 rows are synthetic support guards, not paper-value passes. A
`PROMOTED` row is never the same as an `ACCEPTED` row, and the package never lets
it read as one.

The full ledger behind this one-screen summary is `gp_m2_acceptance()`, a
`gp_m2_acceptance` object whose slots include `table` (the row-by-row gate ledger
with paper and replicated values), `checks` (an alias of `table`),
`effective_registry` (the registry after L02 promotion), `fixture_gates` (the
race/gender fixture evidence used), and `summary` (the counts by status). The
status verb above is the read; the acceptance object is the receipt.

# Part II — Calibration

## 6. The KRW Monte-Carlo study (Table E1)

We now turn from the industry extension to a different kind of question:
*does the grading method's statistical guarantee survive having to estimate the
prior?* KRW answer it with an Online-Appendix Monte-Carlo study — **Table E1** —
and gradepath ships a small stand-in for it. We summarize the study here and
explain the stand-in next.

Table E1 is a **calibration study, not a replication target.** Its logic,
summarized:

- **Known data-generating process, calibrated to the fit.** The estimated prior
  $G$ and the precision-dependence exponent $\beta$ from the *real* deconvolution
  are treated as the simulation's ground truth. Each draw samples firm-level
  random effects from that prior, builds true effects $\theta_i = s_i^{\beta} v_i$
  with the observed standard errors $s_i$ held fixed, and adds normal measurement
  error $\hat\theta_i = \theta_i + s_i\,\varepsilon_i$.
- **Oracle versus empirical.** Two grading rules run on every synthetic dataset.
  The **oracle** uses the true prior that generated the data — it never has to
  estimate anything. The **empirical** re-estimates the prior from the synthetic
  draw through the same pipeline a researcher would actually run. The gap between
  the two gradings is **regret**: the cost of not knowing the prior.
- **What it certifies.** Small regret across draws means estimating the prior
  barely distorts the grades — credible-interval coverage and discordance-rate
  control hold up under prior-estimation uncertainty. This is a statement about
  the *method's* frequentist properties, scoped to the race dimension.

In gradepath's verification substrate these correspond to a block of Table-E1
rows that sit in the published-value registry but are explicitly *outside* the
package's headline scorecard. Only the rendered `tableE1.tex` ships; there is no
machine-readable numeric artifact behind those rows. That fact drives the next
section.

## 7. Why Table E1 is deferred (GP-DEC-18-A)

The deferral of Table E1 from the core replication suite is a deliberate, recorded
design decision — **GP-DEC-18-A**. gradepath's harness checks replicated
quantities against a registry of published values with per-id tolerances. The E1
rows resist that machinery on three independent axes, each sufficient on its own:

1. **No parseable artifact to check against.** The companion ships formatted
   LaTeX (`tableE1.tex`) and nothing machine-readable behind it; the companion
   itself excludes E1 from its own headline suite for exactly this reason. A
   check that cannot read its own ground truth is not a check.
2. **RNG parity is impossible by construction.** The study's posterior
   integration draws from a $10^7$-sample Matlab stream seeded at `15238`.
   gradepath replaces simulation with deterministic quadrature and cannot
   reproduce that stream bit-for-bit, so any reproduction of E1 would be
   approximate by construction — it could match the qualitative conclusion but
   never the exact E1 cell to a tight band.
3. **It validates the method, not a headline number gradepath must hit.** E1
   certifies a *property* — that regret is small and that discordance-rate control
   survives prior estimation — not a published number the package is obligated to
   reproduce. Its cost-to-value for parity is low against the grade counts,
   distributions, and discordance / agreement / fit scalars that *are* in the
   headline suite.

So the full Table E1 Monte-Carlo stays outside the core replication suite, and the
package ships a smaller object that honors the *spirit* of E1 — confirming the
method's guarantees hold — without pretending to a row-for-row fidelity it cannot
deliver. Were E1 to be graduated into the suite some day, two artifacts not
present today would be needed: a `tableE1.tex` parser that recovers its numeric
cells into machine-readable truth, and a Monte-Carlo-aware tolerance class that
compares against a simulation band rather than a point. Until both exist, E1 stays
documented-but-not-checked, and the stand-in below stands in.

## 8. The `gp_calibrate()` stand-in

`gp_calibrate()` is that stand-in: a standalone, seeded, qualitative Monte-Carlo
that runs on a *fitted* model. It treats the fitted prior as the truth, simulates
`n_sim` synthetic datasets from it, re-runs the full one-level pipeline on each,
and reports — as a compact `gp_calibration` object — whether the method's own
frequentist guarantees hold qualitatively. The signature:

```{r calibrate-usage, eval = FALSE}
# Runs one grade-IP solve PER DRAW -- shown, not run. Never runs on render.
fit <- gp_parity_fit("race")
cal <- gp_calibrate(fit,
                    n_sim     = 200L,    # synthetic datasets to draw and re-fit
                    seed      = 15238L,  # in-package reproducibility only
                    dr_target = 0.05,    # the discordance-rate target for dr_ok
                    ci_level  = 0.9,     # the nominal interval level for coverage_ok
                    min_ok    = 1L)      # minimum successful draws before reporting
cal
summary(cal)
```

The return `gp_calibration` carries the run scalars `dr_mean` (mean empirical
discordance rate over successful draws), `coverage` (mean interval coverage), and
`regret_mean` (mean regret of the empirical grading against the oracle), plus the
two boolean verdicts `dr_ok` and `coverage_ok`. It prints as a compact,
result-first one-liner and holds no posterior matrices and no per-draw archive.

How to read the three quantities:

- `dr_mean` near `dr_target` with `dr_ok = TRUE`: the rule's discordance-rate
  control survives re-estimating the prior.
- `coverage` near `ci_level` with `coverage_ok = TRUE`: credible intervals cover
  the truth near nominal under prior-estimation uncertainty.
- `regret_mean` small and non-negative: the empirical grading stays close to the
  oracle that knows the true prior.

**Two scale subtleties, stated plainly,** because they are what make the
quantities mean what they say. First, the oracle grades the *known* true-prior
pairwise **without re-deconvolving** — its pairwise comes straight from the true
prior — and **both** gradings, oracle and empirical, are scored on that *same*
true-prior pairwise. Because the oracle is by definition the risk-minimizer on
that pairwise, the empirical grading can only match or fall short of it, so

$$
\text{regret} \;\ge\; 0 \qquad \text{by construction.}
$$

Regret is never negative; a regret of zero means the re-estimated prior reproduced
the oracle grading exactly. Second, **coverage is assessed on the
$\beta$-invariant $\theta$-scale**: the refit's $\theta$-scale reporting interval
is compared against the true $\theta_i = s_i^{\beta} v_i$. Scoring on the
$\theta$-scale keeps coverage coherent even when a refit's estimated $\beta$
differs from the truth's, which it generally does.

**Small-fit honesty.** The two verdicts are honest, not cosmetic. On a tiny or
weakly identified fit the one-level $\beta$-GMM is noisy and its standardized scale
is ill-identified, so the verdicts are frequently `FALSE` and many synthetic
refits are skipped and counted in `n_failed` (the reported means are over the
successful draws only; if too few succeed the harness errors rather than averaging
noise). Meaningful calibration needs a well-identified fit — KRW's real multi-firm
input. The harness is robust to the occasional non-convergent draw and reports
honestly either way; it never tunes the data to force a pass.

To keep the boundary sharp, `gp_calibrate()` deliberately is **not** seeded to
match Matlab's `15238` stream (the shared default seed is a cosmetic homage;
gradepath's RNG and its quadrature substitute mean the synthetic data differ from
the study by construction), **not** the full multi-model, multi-hundred-draw grid
of the study (it runs a small `n_sim` on the already-fitted model), and **not** a
member of any default scorecard, headline result, or reproduction of any Table E1
cell. It is a diagnostic. Because each draw re-solves a grade integer program, it
**never runs on package load or on vignette render** — every calibration chunk
here is `eval = FALSE` — and it prints a one-line budget heads-up before the loop
when `n_sim` times the unit count exceeds 600, an informational message that never
changes the result.

## 9. Where to next

That closes the method track. Over five vignettes the one-level pipeline was
derived and verified end to end, the grading was extended to the two-level
industry model and reported honestly as `PARTIAL_ACCEPTED`, and the method's
frequentist guarantees were given a seeded stand-in for the deferred Table E1
study. To revisit any stage:

- **Foundations and notation** — `r vlink("m1-foundations-and-notation", "m1")`
  fixes the symbol table, the grade-direction triple, and the firms-vs-names
  welfare flip every later vignette relies on.
- **Precision and standardization** —
  `r vlink("m2-precision-and-standardization", "m2")` derives the r-scale and the
  $\beta$-GMM step, the same precision dependence the two-level model carries.
- **Deconvolution and posterior** —
  `r vlink("m3-deconvolution-and-posterior", "m3")` builds the one-level prior,
  the posteriors, and the pairwise matrix $\Pi$ this vignette extends to two
  levels and re-uses under calibration.
- **Grading, the frontier, and report cards** —
  `r vlink("m4-grading-frontier-and-report-cards", "m4")` derives the grade
  integer program and the frontier that both the two-level routes and each
  calibration draw solve.

For the workflow side, the applied track runs the same pipeline hands-on:
`r vlink("a1-getting-started", "a1")` is the first contact,
`r vlink("a2-the-grading-workflow", "a2")` walks the grading stage with race and
gender side by side, `r vlink("a3-reading-and-exporting-results", "a3")` covers
the report-card and frontier accessors, `r vlink("a4-figures", "a4")` is the
figure gallery, and `r vlink("a5-solvers-and-calibration", "a5")` covers the
solver backends and the calibration verbs in workflow terms. Throughout, the grade
is a sorting by posterior evidence, never a contest: gradepath's printed output is
result-first and quantity-anchored (the package decision `GP-DEC-14-A`), so the
same code path serves firms and names — and the two-level routes and the
calibration draws — without the printed text inverting on one of them. The grade
label is an integer in $\{1, \dots, n\}$, a sorting index, not a letter grade.

### Provenance

```{r provenance}
sessionInfo()
```

## References
